---
title: "Clustered bar charts"
format: html
editor: visual
---

```{r setup}
#| message: false
library(tidyverse)
library(visu)
```

### How bar_clustered() works

Same idea as `bar()`, but for two categorical variables at once: `x` groups bars along the axis, `fill` groups bars side by side within each `x` group (`position = "dodge2"`). Colours come from `nc_alt1` via `scale_fill_pal_categorical()`, sized automatically to `fill`'s number of levels — no `fill` colour argument to set directly, unlike `bar()`.

-   `dat` — the data frame, one row per x/fill combination
-   `x` — unquoted column name for the axis grouping
-   `y` — unquoted column name for the already-computed value each bar shows
-   `fill` — unquoted column name for the within-group colour grouping
-   `x_lab` / `y_lab` / `fill_lab` — optional axis/legend titles
-   `title` — optional plot title
-   `base_size` — base font size, via `theme_house()`
-   `horiz` — `FALSE` (default) for upright bars, `TRUE` to flip horizontal - done by swapping which column maps to the x/y aesthetic, not `coord_flip()` (see `9_basic_bar_charts.qmd`)
-   `label_digits` — decimal places on each bar's value label, default 1
-   `label_size` — font size for value labels, default `NULL` (falls back to `base_size`), independent of it
-   `label_style` — `"normal"` (default), `"suppress"` (no value labels), or `"diag"` (labels at 45 degrees)
-   `fill_family` — `NULL` (default) colours `fill` categorically, spread across `nc_alt1` families (`scale_fill_pal_categorical()`); set to a family name (e.g. `"mint"`) to colour it sequentially within that one family instead (`scale_fill_pal_sequential()`) - suited to an ordered `fill` variable, no cap on levels
-   `plot_width` / `plot_height` — feed label-wrapping maths only, not the plot's actual rendered size (see `9_basic_bar_charts.qmd` for the fuller explanation)

`diamonds` `cut`/`color` rather than `2_geom_gallery.qmd`'s `class`/`drv` — that pairing is too sparse to show clustering properly (several classes only have one drivetrain in the data at all, so most "clusters" are really just one bar). Every `cut`/`color` combination is well populated, so each cluster actually shows the full comparison.

```{r}
diamonds_cut_colour <- diamonds %>% count(cut, color)

bar_clustered(diamonds_cut_colour, cut, n, color, title = "Diamonds by cut and colour", fill_lab = "Colour")
```

### The overlap problem, and fixing it

With 7 `color` levels dodged inside each `cut` group, the bars get narrow but the value labels stay at `base_size` regardless, so the numbers collide with their neighbours - visible above. `label_size` and `label_digits` are the direct fix: shrinking the label font gives it room to fit, and dropping the decimal (`n` is a whole-number count, so ".0" is pure wasted width) shortens the text on top of that.

```{r}
bar_clustered(diamonds_cut_colour, cut, n, color, title = "Diamonds by cut and colour", fill_lab = "Colour", label_size = 6, label_digits = 0)
```

If that's still too tight, `label_style = "suppress"` drops the value labels altogether and leans on the axis/legend instead - a reasonable call once a chart is dense enough that no label size survives gracefully.

```{r}
bar_clustered(diamonds_cut_colour, cut, n, color, title = "Diamonds by cut and colour", fill_lab = "Colour", label_style = "suppress")
```

### horiz

Flips to horizontal bars - each `cut` category gets its own row, so the 7 dodged `color` bars within it have more room top-to-bottom rather than competing for width.

```{r}
bar_clustered(diamonds_cut_colour, cut, n, color, title = "Diamonds by cut and colour", fill_lab = "Colour", horiz = TRUE, label_size = 6, label_digits = 0)
```

### label_style = "diag"

Rotates the value labels 45 degrees instead - an alternative to shrinking them, for the upright case.

```{r}
bar_clustered(diamonds_cut_colour, cut, n, color, title = "Diamonds by cut and colour", fill_lab = "Colour", label_style = "diag", label_digits = 0, label_size = 8)
```

### fill_family

`color` is itself ordered (D best through J worst), so a one-hue sequential ramp (`scale_fill_pal_sequential()` under the hood) carries that ordering in the colour itself, rather than spreading across families the way the categorical default does. 7 levels is past the 5 discrete shades one family has, so this draws on the interpolated branch.

```{r}
bar_clustered(diamonds_cut_colour, cut, n, color, title = "Diamonds by cut and colour", fill_lab = "Colour", fill_family = "purple", label_size = 6, label_digits = 0)
```
