---
title: "Stacked bar charts"
format: html
editor: visual
---

```{r setup}
#| message: false
library(tidyverse)
library(visu)
```

### How bar_stacked() works

Same two-variable shape as `bar_clustered()`, but stacks segments into one bar per `x` category (`position = "stack"`) instead of placing them side by side — suited to more `x` categories, or where the total across `fill` levels matters as much as each individual segment.

-   `dat` — the data frame, one row per x/fill combination
-   `x` — unquoted column name for the axis grouping
-   `y` — unquoted column name for the already-computed value each bar shows
-   `fill` — unquoted column name for the stacked colour grouping
-   `x_lab` / `y_lab` / `fill_lab` — optional axis/legend titles
-   `title` — optional plot title
-   `base_size` — base font size, via `theme_house()`
-   `horiz` — `TRUE` (default, matching this function's original always-flipped behaviour) for horizontal bars, `FALSE` for upright - done by swapping which column maps to the x/y aesthetic, not `coord_flip()` (see `9_basic_bar_charts.qmd`)
-   `label_digits` — decimal places on each segment's value label, default 1
-   `label_size` — font size for value labels, default `NULL` (falls back to `base_size`), independent of it
-   `label_style` — `"normal"` (default), `"suppress"` (no value labels), or `"diag"` (labels at 45 degrees)
-   `fill_family` — `NULL` (default) colours `fill` categorically, spread across `nc_alt1` families (`scale_fill_pal_categorical()`); set to a family name (e.g. `"mint"`) to colour it sequentially within that one family instead (`scale_fill_pal_sequential()`) - suited to an ordered `fill` variable, no cap on levels
-   `legend_position` — `"bottom"` (default) or `"right"`
-   `plot_width` / `plot_height` — feed label-wrapping maths only, not the plot's actual rendered size (see `9_basic_bar_charts.qmd` for the fuller explanation)

Same `clarity`/`cut` data as `6_house_palette.qmd`'s stacked bar check — flagged there as the harder legibility test, since neighbouring swatches sit right next to each other inside one bar. Also a better fit for `bar_stacked()` than a smaller grouping: 8 `clarity` levels justify the horizontal flip.

```{r}
diamonds_clarity_cut <- diamonds %>% count(clarity, cut)

bar_stacked(diamonds_clarity_cut, clarity, n, cut, title = "Diamonds by clarity and cut", fill_lab = "Cut")
```

### label_digits, label_size, label_style

Same three label arguments as `bar()`/`bar_clustered()`. `n` is a whole-number count, so `label_digits = 0` drops the wasted ".0"; stacked segments are usually roomier than dodged bars, but the same `label_size`/`"suppress"` options are there if a particular chart gets tight.

```{r}
bar_stacked(diamonds_clarity_cut, clarity, n, cut, title = "Diamonds by clarity and cut", fill_lab = "Cut", label_style = "suppress")
```

### fill_family

`cut` is itself ordered (Fair worst through Ideal best), so a one-hue sequential ramp carries that ordering in the colour rather than spreading across families the way the categorical default does.

```{r}
bar_stacked(diamonds_clarity_cut, clarity, n, cut, title = "Diamonds by clarity and cut", fill_lab = "Cut", fill_family = "mint")
```

### horiz = FALSE

Previously impossible - `bar_stacked()` always flipped via a hardcoded `coord_flip()`. Now it's just the default being overridden, for an upright stacked bar instead.

```{r}
bar_stacked(diamonds_clarity_cut, clarity, n, cut, title = "Diamonds by clarity and cut", fill_lab = "Cut", 
            horiz = FALSE, label_style = "suppress")
```

`label_style = "diag"` on the same upright chart - rotates the segment labels 45 degrees.

```{r}
bar_stacked(diamonds_clarity_cut, clarity, n, cut, title = "Diamonds by clarity and cut", fill_lab = "Cut", horiz = FALSE, label_digits = 0, label_style = "diag")
```

<br>

```{r}
diamonds %>% count(cut, color) %>% group_by(cut) %>%
  mutate(p = n / sum(n) * 100) %>%
  bar_stacked(cut, p, color, y_lab = "Colour (%)", x_lab = "Cut quality", 
              title = "Colour by cut quality") 

```

<br> 

```{r}

diamonds %>% mutate(
  price2 = cut(price, breaks = 5, 
               labels = c("326-4,025","4,026-7,724","7,725-11,422","11,427-15,122","15,124-18,823"))) %>% 
  # group_by(price2) %>% summarise(min = min(price), max = max(price))
  count(cut, price2) %>% group_by(cut) %>% mutate(cut = fct_rev(cut)) %>%
  mutate(p = n / sum(n) * 100) %>% 
  # select(-n) %>% pivot_wider(values_from = p, names_from = price2)
  bar_stacked(x = cut, y = p, fill = price2, label_style = "suppress", fill_family = "orange")
```
















