---
title: "ggplot intro"
format: html
editor: visual
---

## The layer skeleton

A ggplot call builds up in layers. ggplot doesn't care what order you write them in — `+` just adds elements to the plot object — but writing them in this order mirrors the underlying grammar and keeps calls readable.

```{r}
#| eval: false

# data + global (inherited) mapping
ggplot(data = df, mapping = aes(x = xvar, y = yvar)) +   
  
  # geom: aes() = mapped, arg = fixed
  geom_col(aes(fill = grp), colour = "black") +         
  # stat = "identity" is geom_col's default; 
  # use geom_bar(stat = "count") to let ggplot tally rows

  # scale: controls how mapped values -> visual controls
  scale_fill_manual(values = my_palette) +      
  scale_y_continuous(labels = scales::comma) +

  # coord: alters the coordinate space (e.g. horizontal bars)
  coord_flip() +                                          

  # facet: splits into small multiples
  facet_wrap(~ facet_var, scales = "free_y") +             

  labs(title = "Title", x = "X axis", y = "Y axis", fill = "Legend") +

  # theme: non-data styling, best applied last
  theme_minimal() +                                       
  theme(legend.position = "bottom")
```

## Setting aesthetics `aes`

The one conceptual distinction worth keeping straight, since it's the actual source of bugs:

**Mapped vs fixed.**

-   `aes(colour = var)` maps a column to a visual property — it varies row to row.\
-   `colour = "black"` outside `aes()` sets it to one fixed value for the whole layer.

**Global vs local.**

-   `aes()` supplied inside `ggplot()` is inherited by every geom that follows.\
-   `aes()` supplied inside a specific `geom_*()` only applies to that layer, and overrides the global mapping for that aesthetic.

<br>

## Presentation vs calculation of statistics

Different ggplot geoms approach this differently. Some calculate new statistics on the data you pass them, others simply plot the data already there.

-   `geom_bar()` uses stat 'count' to count the number of cases in x category,\
-   `geom_col()` uses stat 'identity'. It plots the values already in `y`; bar height = `y`,\
-   `geom_boxplot()` does much more, it calculates IQR, median, max and min, and presents them all,\
-   `geom_histogram()` bins continuous variables and counts the observations per bin.

These are controlled by the `stat` argument, which defines the calculation used.

<br>

## Applied examples

Worked against `mpg` and `economics_long`, both built into ggplot2 — no external data needed to reproduce any of these.

| Concept | Illustrated with |
|---|---|
| x / y axis | `geom_point` |
| Colour vs fill | `geom_col` |
| Stat: calculated vs pre-supplied | `geom_bar` → `geom_col` → `geom_histogram` |
| Position adjustments | `geom_col` |
| Axis reordering | `geom_col` |
| Continuous vs discrete colours | `geom_point` |
| Size | `geom_point` |
| Shape | `geom_point` |
| Alpha (transparency) | `geom_point` |
| Linetype | `geom_line` |
| Group | `geom_line` (+ spaghetti plot) |
| Data labels | `geom_col` + `geom_text` |
| Facets | `geom_boxplot` |
| Overlaying geoms | `geom_boxplot` + `geom_jitter` |

```{r setup}
#| message: false
library(tidyverse)
```

### x / y axis

The two positional aesthetics. Both continuous here. Each point's screen position is read straight off the data.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

### Colour vs fill

`colour` is the outline, `fill` the fill colour.

```{r}
mpg_class <- mpg %>% count(class)

ggplot(mpg_class, aes(x = class, y = n)) +
  # fill = bar interior, colour = bar outline — both visible at once
  geom_col(fill = "steelblue", colour = "black", linewidth = 0.8) +
  coord_flip()
```

### Stat: calculated vs pre-supplied

Below we have the same chart, arrived at in two ways.

`geom_bar` takes raw rows and counts them; `geom_col` takes a value that's already been counted.

```{r}
# geom_bar: raw rows in, ggplot counts them (stat = "count")
ggplot(mpg, aes(x = class)) +
  geom_bar(fill = "steelblue", colour = "black")
```

Here stat is created separately using count(), then presented with geom_col().

```{r}
# geom_col: pre-aggregated, ggplot just plots n (stat = "identity")
mpg_class <- mpg %>% count(class)
ggplot(mpg_class, aes(x = class, y = n)) +
  geom_col(fill = "steelblue", colour = "black")
```

There are lots of statistics ggplot can calculate on the back end, for example, a third stat type — `geom_histogram` bins a continuous variable, then counts rows per bin (stat = "bin"):

```{r}
ggplot(mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2, fill = "steelblue", colour = "black")
```

### Position adjustments

Same grouped data, three ways to arrange the `fill` segments within each bar.

```{r}
mpg_class_drv <- mpg %>% count(class, drv)

# stack (default): segments sit on top of each other
ggplot(mpg_class_drv, aes(x = class, y = n, fill = drv)) +
  geom_col(position = "stack") +
  coord_flip()

# dodge: segments sit side by side
ggplot(mpg_class_drv, aes(x = class, y = n, fill = drv)) +
  geom_col(position = "dodge") +
  coord_flip()

# fill: stacked and rescaled to proportions (0-1) — good for share-of-total
ggplot(mpg_class_drv, aes(x = class, y = n, fill = drv)) +
  geom_col(position = "fill") +
  coord_flip()
```

### Axis reordering

Discrete axes default to factor/alphabetical order, which is rarely what you want on a ranked bar chart.

```{r}
# default: alphabetical
ggplot(mpg_class, aes(x = class, y = n)) +
  geom_col() +
  coord_flip()

# reordered by value
ggplot(mpg_class, aes(x = fct_reorder(class, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(x = "class")
```

### Continuous vs discrete colours

Same aesthetic (`colour`), different variable type — the scale and legend change accordingly.

```{r}
# continuous -> gradient legend
ggplot(mpg, aes(x = displ, y = hwy, colour = cty)) +
  geom_point()

# discrete -> key legend
ggplot(mpg, aes(x = displ, y = hwy, colour = class)) +
  geom_point()
```

### Size

```{r}
ggplot(mpg, aes(x = displ, y = hwy, size = cty)) +
  geom_point(alpha = 0.6)
```

### Shape

Discrete only, point geoms only — and unreadable much past 5-6 levels. `drv` has three.

```{r}
ggplot(mpg, aes(x = displ, y = hwy, shape = drv)) +
  geom_point(size = 2)
```

### Alpha (transparency)

Fixed here rather than mapped — every point equally transparent, which is what you want for overplotting.

Some appear darker here - this is because some points overlap. This is what `alpha` is telling you. 

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.3, size = 3)
```

<br>

Here you can see the overlap increasing opacity. 

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_jitter(alpha = 0.3, size = 3)
```

### Linetype

```{r}
econ_subset <- economics_long %>% filter(variable %in% c("psavert", "uempmed"))

ggplot(econ_subset, aes(x = date, y = value01, linetype = variable)) +
  geom_line()
```

### Group

The classic bug: with no colour or group mapping, ggplot draws one line straight through every row in row order not one line per series.

```{r}
# broken: one zigzag line through all five series
ggplot(economics_long, aes(x = date, y = value01)) +
  geom_line()

# fixed: group (or colour) tells ggplot to draw one line per series
ggplot(economics_long, aes(x = date, y = value01, group = variable)) +
  geom_line()
```

**Spaghetti plot — many individual trajectories, no colour.** The standard diagnostic for repeated-measures/panel data before fitting a mixed model: one line per respondent, `group = id`, deliberately not mapped to colour (colouring 30+ respondents just produces an unreadable legend).

```{r}
set.seed(1)
panel_data <- tibble(
  id   = rep(1:30, each = 5),
  wave = rep(1:5, times = 30)
) %>%
  group_by(id) %>%
  mutate(score = 50 + (wave - 1) * rnorm(1, 2, 1) + rnorm(1, 0, 8) + rnorm(5, 0, 4)) %>%
  ungroup()

ggplot(panel_data, aes(x = wave, y = score, group = id)) +
  geom_line(alpha = 0.3)
```

### Data labels

Putting the value directly on the chart — `geom_text()` (or `geom_label()` for a background box) layered on top of another geom.

```{r}
ggplot(mpg_class, aes(x = class, y = n)) +
  geom_col(fill = "steelblue", colour = "black") +
  geom_text(aes(label = n), vjust = -0.3) +
  coord_cartesian(clip = "off")
```

Labelling individual scatter points is a different problem — points overlap, so a bare `geom_text()` produces unreadable clutter. The standard fix is the `ggrepel` package (`geom_text_repel()` / `geom_label_repel()`), which nudges overlapping labels apart automatically. Not demonstrated here since it needs an extra package, but it's the first thing to reach for the moment point labels overlap.

### Facets

```{r}
# facet_wrap: one variable, wrapped into a grid
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot() +
  facet_wrap(~ drv)

# facet_grid: two variables, strict rows x columns
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot() +
  facet_grid(drv ~ year)
```

### Overlaying geoms

Layers stack — nothing stops you combining two geoms on the same data to get the benefit of both. Here, the boxplot's summary shape plus every individual observation from `geom_jitter`. `outlier.shape = NA` turns off the boxplot's own outlier points so they don't get drawn twice (once by the boxplot, once by jitter).

```{r}
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, colour = "blue")
```
