---
title: "Geom gallery"
format: html
editor: visual
---

A single, basic worked example of each geom on the original "core-ish" list — standalone from the intro chapter, so nothing here depends on objects created there. Worked against `mpg` and `economics`, both built into ggplot2, plus `ggrepel` for the two repel geoms (`install.packages("ggrepel")` if not already installed).

```{r setup}
#| message: false
library(tidyverse)
library(ggrepel)
```

### geom_point

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

### geom_line

```{r}
ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line()
```

### geom_col / geom_bar

`geom_bar` counts raw rows for you (stat `"count"`); `geom_col` plots a value you've already computed (stat `"identity"`).

```{r}
ggplot(mpg, aes(x = class)) +
  geom_bar(fill = "steelblue", colour = "black")

mpg_class <- mpg %>% count(class)

ggplot(mpg_class, aes(x = class, y = n)) +
  geom_col(fill = "steelblue", colour = "black")
```

### Grouped bar chart

`position = "dodge"` — one bar per drivetrain, side by side within each class. Flipped horizontal here (`coord_flip()`), which is the usual fix once category labels get too long to sit legibly on the x axis.

```{r}
mpg_class_drv <- mpg %>% count(class, drv)

ggplot(mpg_class_drv, aes(x = class, y = n, fill = drv)) +
  geom_col(position = "dodge") +
  coord_flip()
```

### Stacked bar chart

`position = "stack"` — ggplot's default for `geom_col`/`geom_bar` when `fill` is mapped, so it doesn't need to be stated explicitly. Same data as the grouped chart above, segments stacked into one bar per class instead of sitting side by side.

```{r}
ggplot(mpg_class_drv, aes(x = class, y = n, fill = drv)) +
  geom_col()
```

### geom_histogram

```{r}
ggplot(mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2, fill = "steelblue", colour = "black")
```

### geom_density

```{r}
ggplot(mpg, aes(x = hwy)) +
  geom_density(fill = "steelblue", alpha = 0.5)
```

### geom_boxplot

```{r}
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot() 
```

<br> 
With an overlay...
```{r}
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.2, colour = "blue")
```

### geom_violin

Same shape of comparison as the boxplot above, but shows the full distribution shape rather than just the five-number summary.

```{r}
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_violin() 
```

### geom_jitter

`cty` vs a discrete x — plain points would stack exactly on top of each other; jitter adds a small amount of random noise so overlapping observations are visible.

```{r}
ggplot(mpg, aes(x = class, y = cty)) +
  geom_jitter(width = 0.2, alpha = 0.5)
```

### geom_smooth

A fitted trend line with its confidence ribbon, layered over the raw points.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm")
```

### geom_errorbar / geom_pointrange

Uncertainty around a point estimate — here, mean `hwy` by class ± 1 standard error.

```{r}
mpg_summary <- mpg %>%
  group_by(class) %>%
  summarise(mean_hwy = mean(hwy), se_hwy = sd(hwy) / sqrt(n()))

ggplot(mpg_summary, aes(x = class, y = mean_hwy)) +
  geom_errorbar(aes(ymin = mean_hwy - se_hwy, ymax = mean_hwy + se_hwy), width = 0.2) +
  geom_point()

# geom_pointrange does the same thing in a single geom
ggplot(mpg_summary, aes(x = class, y = mean_hwy)) +
  geom_pointrange(aes(ymin = mean_hwy - se_hwy, ymax = mean_hwy + se_hwy))
```

### geom_area / geom_ribbon

`geom_area` fills the space under a single line; `geom_ribbon` fills the space between two values (`ymin`/`ymax`) — here an illustrative ±10% band around the same series.

```{r}
ggplot(economics, aes(x = date, y = unemploy)) +
  geom_area(fill = "steelblue", alpha = 0.6)

ggplot(economics, aes(x = date)) +
  geom_ribbon(aes(ymin = unemploy * 0.9, ymax = unemploy * 1.1), fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = unemploy))
```

### geom_text / geom_label

Value labels placed directly on a chart — `geom_label` adds a background box, `geom_text` doesn't.

```{r}
ggplot(mpg_class, aes(x = class, y = n)) +
  geom_col(fill = "steelblue", colour = "black") +
  geom_text(aes(label = n), vjust = -0.3)

ggplot(mpg_class, aes(x = class, y = n)) +
  geom_col(fill = "steelblue", colour = "black") +
  geom_label(aes(label = n), vjust = -0.3)
```


### geom_text_repel / geom_label_repel (ggrepel)

Plain `geom_text` on a scatter plot produces overlapping, unreadable labels. `ggrepel`'s versions nudge labels apart automatically while keeping a leader line back to the point.

```{r}
mpg_class_means <- mpg %>%
  group_by(class) %>%
  summarise(displ = mean(displ), hwy = mean(hwy))

ggplot(mpg_class_means, aes(x = displ, y = hwy, label = class)) +
  geom_point() +
  geom_text_repel()
```


### geom_tile

The actual case for reaching for this: a correlation matrix. A table of numbers is slow to scan for pattern; colour lets your eye pick out the strong relationships immediately, and overlaying the coefficient keeps the precision a table gives you. Same logic applies to any variable-by-variable or category-by-category grid — cross-tab percentages, a similarity matrix, a confusion matrix.

```{r}
mtcars_cor <- mtcars %>%
  select(mpg, disp, hp, drat, wt, qsec) %>%
  cor() %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation")

ggplot(mtcars_cor, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), size = 3) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick", midpoint = 0) +
  labs(x = NULL, y = NULL)
```

