---
title: "Layering a plot, stage by stage"
format: html
editor: visual
---

The skeleton in the intro chapter shows the full anatomy of a ggplot call at once. This chapter rebuilds it one layer at a time on a single running example — `diamonds` (built into ggplot2) — so the effect of each addition is visible on its own before the next one goes on.

**The question:** how does the mix of cut quality within each clarity grade change, and does that composition differ by colour grade? This is a *share* question, not a volume one — and that shapes several of the choices made from stage 2 onward.

### The data

`diamonds` is \~54,000 round-cut diamonds with price and nine physical/quality attributes. Three of those are the ones in play here, and all three are graded, ordered categories rather than arbitrary labels:

-   **`cut`** — quality of the cut: Fair (worst), Good, Very Good, Premium, Ideal (best)
-   **`color`** — diamond colour grade: D (best, colourless) through J (worst, most visible tint)
-   **`clarity`** — how free of inclusions the stone is: I1 (worst) through IF (best, internally flawless)

None of these are things you'd normally sum or average directly — they're categorical groupings, which is exactly the shape of variable a bar chart is for. `diamonds_summary` collapses the \~54,000 individual diamonds down to one row per combination of the three, with `n` giving the count of diamonds in each — that's what gets plotted throughout, not the raw data.

Neither `clarity` nor `color` reads as obviously ordered from its labels alone — nothing about "I1" or "D" tells a viewer which end is best. Rather than fixing that with scale/labeller code at every plotting stage, it's simplest to relabel the two extreme levels once, in the data, before any plot is built — `fct_recode()` renames a factor level without disturbing its position in the level order, so every chart from here on inherits the tagged labels automatically.

```{r setup}
#| message: false
library(tidyverse)
library(scales)
```

```{r}
diamonds_summary <- diamonds %>%
  count(clarity, cut, color) %>%
  mutate(
    clarity = fct_recode(clarity, "I1 (worst)" = "I1", "IF (best)" = "IF"),
    color   = fct_recode(color,   "D (best)" = "D",   "J (worst)" = "J")
  )
```

### 1. Create plot

Just `x` and `y` mapped, default geom settings. With no `fill` mapped, `geom_col`'s default `position = "stack"` silently collapses the `cut`/`color` breakdown into one bar per clarity grade — this is the total count per clarity. It's a real number, but it doesn't answer the actual question: it says nothing about how the mix of cut quality shifts from one clarity grade to the next, since clarity grades have wildly different total volumes to begin with.

```{r}
plot <- ggplot(diamonds_summary, aes(x = clarity, y = n)) +
  geom_col()

plot
```

### 2. fill colour

`fill = cut` is mapped — it varies by row, splitting each bar into its `cut` segments. But comparing segment *sizes* across bars only works if every bar is the same total height — otherwise a clarity grade with more diamonds overall will always look like it has "more" of every cut, regardless of the actual mix. `position = "fill"` rescales every bar to 100%, so segment proportions become directly comparable across clarity grades.

Confusingly, "fill" means two different things in this one call: `aes(fill = cut)` is the *aesthetic* being coloured by; `position = "fill"` is the *position adjustment* that rescales and stacks. Same word, unrelated arguments.

```{r}
plot <- ggplot(diamonds_summary, aes(x = clarity, y = n, fill = cut)) +
  geom_col(position = "fill")

plot

```

The composition story is already visible here in a way stage 1 couldn't show: better clarity grades visibly carry a larger share of Ideal/Premium cuts.

### 3. Colour palette

Swapping ggplot's default hue palette for a deliberate one.

```{r}
cut_palette <- c(
  "Fair"      = "#b2182b",
  "Good"      = "#ef8a62",
  "Very Good" = "#fddbc7",
  "Premium"   = "#67a9cf",
  "Ideal"     = "#2166ac"
)

plot <- plot +  scale_fill_manual(values = cut_palette)
plot

```

### 4. Axis scale

The y-axis is a proportion now (0 to 1), not a raw count — `scales::comma` is the right tool for large counts, but it's the wrong one here. `scales::percent` matches what the axis actually represents.

```{r}
plot <- plot +  scale_y_continuous(labels = percent)
plot
```

### 5. Axis flip

`clarity`'s labels are short enough to survive on a vertical axis here, but this is the point in the build where you'd reach for `coord_flip()` if they weren't — horizontal bars, category labels read left to right instead of rotated.

```{r}
plot <- plot +  coord_flip()
plot
```

### 6. Add facets

Splitting into one panel per colour grade answers the second half of the original question directly: does the cut composition pattern seen above hold across colour grades, or does it shift? The "(best)"/"(worst)" tags on the two extreme colour grades — baked into the data up front — carry straight through to the facet strip labels here with no extra code.

```{r}
plot <- plot + facet_wrap(~ color, scales = "free_x")
plot 
```

The gotcha worth flagging separately: after `coord_flip()`, `free_x`/`free_y` in `facet_wrap()` swap meaning. Facet scales are evaluated on the physical, post-flip axes — not on the original `x`/`y` aesthetic roles the way theme elements are. Since the share axis (`n`, rescaled by `position = "fill"`) now renders horizontally, it's `scales = "free_x"` that frees it per panel, not `"free_y"` — easy to get backwards.

### 7. Label plot

```{r}
plot <- plot +
  labs(
    title = "Cut composition by clarity, faceted by colour grade",
    x     = "Clarity",
    y     = "Share of diamonds",
    fill  = "Cut"
  )

plot
```

### 8. Theme

Non-data styling, applied last, in three passes rather than one jump.

**8a — base theme.** Swap ggplot's default grey panel for a plain one.

```{r}
plot <- plot +  theme_minimal()
plot
```

**8b — legend placement and sizing.** Seven facets and five cut categories already crowd the plot — move the legend below and shrink the keys so it doesn't compete with the panels.

```{r}
plot <- plot  +
  theme(
    legend.position = "bottom",
    legend.key.size = unit(0.4, "cm")
  )

plot
```

**8c — text sizing and gridline cleanup.** The export-ready pass: consistent text sizes across every text element, minor gridlines dropped, facet strip labels tidied.

```{r}
plot <- plot + 
  theme(
    legend.position   = "bottom",
    legend.key.size   = unit(0.4, "cm"),
    legend.text       = element_text(size = 8),
    legend.title      = element_text(size = 9),
    axis.text         = element_text(size = 8, colour = "black"),
    axis.title        = element_text(size = 10),
    plot.title        = element_text(size = 12, face = "bold", hjust = 0.5),
    strip.text        = element_text(size = 8, face = "bold"),
    panel.grid.minor  = element_blank()
  )

plot
```

### Putting it all together

The same chart, all stages combined into the one call you'd actually write:

```{r}
ggplot(diamonds_summary, aes(x = clarity, y = n, fill = cut)) +

  # geom: fill mapped (varies by cut). position = "fill" rescales each bar
  # to 100% so cut composition is comparable across clarity grades,
  # regardless of their total volume
  geom_col(position = "fill") +

  # scale: deliberate palette + percent-formatted shares (not comma —
  # this axis is a proportion, not a count)
  scale_fill_manual(values = cut_palette) +
  scale_y_continuous(labels = percent) +

  # coord: horizontal bars
  coord_flip() +

  # facet: one panel per colour grade — free_x, not free_y, since coord_flip
  # already swapped which physical axis carries the share
  facet_wrap(~ color, scales = "free_x") +

  labs(
    title = "Cut composition by clarity, faceted by colour grade",
    x     = "Clarity",
    y     = "Share of diamonds",
    fill  = "Cut"
  ) +

  # theme: base, then legend, then the export-ready text/gridline pass
  theme_minimal() +
  theme(
    legend.position   = "bottom",
    legend.key.size   = unit(0.4, "cm"),
    legend.text       = element_text(size = 8),
    legend.title      = element_text(size = 9),
    axis.text         = element_text(size = 8, colour = "black"),
    axis.title        = element_text(size = 10),
    plot.title        = element_text(size = 12, face = "bold", hjust = 0.5),
    strip.text        = element_text(size = 8, face = "bold"),
    panel.grid.minor  = element_blank()
  )
```
