---
title: "ggplot themes"
format: html
editor: visual
---

#### Load some pacakges   

```{r, warning=FALSE,message=FALSE}
library(tidyverse)
library(flextable)
library(ggpubr)
library(kableExtra)
```

<br>

#### Load the mpg dataset   

```{r}
data(iris)
```

<br>

Every row in mpg is a car, every column a feature of that car. mpg stands for miles per gallon. 

```{r}
head(iris)
```

## Standard themes    

The default theme, featuring a gray background and white gridlines.

### theme_gray

```{r}
plot <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point()  

plot + theme_gray()
```

<br>

### theme_bw  

A black on white theme.

```{r}
plot + theme_bw()
```

<br>

### theme_linedraw

A theme with only black lines of various widths on white backgrounds.

```{r}
plot + theme_linedraw()
```

<br>

### theme_light

A theme similar to theme_linedraw but with grey lines and axes designed to draw more attention to the data.

```{r}
plot + theme_light()
```

<br>

### theme_dark

A theme similar to theme_light, but with a dark background. A useful theme for making thin colored lines stand out.

```{r}
plot + theme_dark()
```

<br>

### theme_minimal

A theme with no background annotations.

```{r}
plot + theme_minimal()
```

<br>

### theme_classic

A theme with no gridlines.

```{r}
plot + theme_classic()
```

<br>

### theme_void

A completely empty theme.

```{r}
plot + theme_void()
```

<br>

## ggthemes library  

Next, we’ll show a few examples of how to use the predefined themes to modify the aesthetics of plots.  

```{r}
library(ggthemes)
```

<br>

### theme_wsj   

The Wall Street Journal's theme.  

```{r}
plot + theme_wsj()
```

<br>

### theme_tufte  

A minimalist theme inspired by the work of statistician Edward Tufte.   

```{r}
plot + theme_tufte()
```

<br>

### theme_solarized  

A theme that uses colors based on the solarized palette.

```{r}
plot + theme_solarized()
```

<br>

```{r}
plot + theme_solarized(light = FALSE)
```

<br>

### theme_gdocs  

A theme with Google Docs Chart defaults.

```{r}
plot + theme_gdocs()
```

<br>

### theme_fivethirtyeight  

Theme inspired by fivethirtyeight.com plots.

```{r}
plot + theme_fivethirtyeight()
```

<br>

### theme_economist  

Theme inspired by The Economist.

```{r}
plot + theme_economist()
```

<br>

## Sources   

[ggplot2 themes](https://www.statology.org/best-ggplot2-themes/)   