---
title: "Bar charts"
format: html
editor: visual
---

#### Set up

```{r, warning=FALSE,message=FALSE}
library(tidyverse)
library(ggpubr)
data(mpg)
```

Every row in mpg is a car, every column a feature of that car. mpg stands for miles per gallon. 

```{r}
head(mpg)
```

<br> 

## Basics of bar charts    

These uses the geom_col() function to display values *in* the data, geom_bar() to count the cases in a group.   
 
In this example, average highway miles per gallon (hwy) is displayed by a the type of car (class).  

```{r}

(hwy_class <- mpg %>% 
  group_by(class) %>% 
  summarise_at(vars(hwy), list(average_hwy= ~round(mean(.x), digits = 1))))

```

<br>

### A basic bar chart

That data is then fed to ggplot to produce a simple bar chart. 

```{r}

ggplot(data=hwy_class) + 
  geom_col(mapping=aes(x=class, y=average_hwy))

```

<br> 

### A horizontal bar chart

```{r}
ggplot(data=hwy_class) +
  geom_col(mapping=aes(x=class, y=average_hwy)) +
  coord_flip()
```

<br>

## Amend fill colour 

<br>

#### Standard gglot colours

```{r}

plot <- ggplot(data=hwy_class) +
  aes(x=class, y=average_hwy, fill=class) +
  geom_col(width=0.5, colour = "black") +
  theme_minimal()

plot

```

#### Other options

```{r}

ggarrange(
plot + scale_fill_grey() + theme(legend.position="none"),
plot + scale_fill_brewer(palette="PuRd") + theme(legend.position="none"),
plot + scale_fill_brewer(palette="Blues") + theme(legend.position="none"),
ncol = 1, nrow = 3)

```

<br>

### Fill by another variable  

```{r}

(hwy_class_year <- mpg %>% 
  group_by(class, year) %>% 
  summarise_at(vars(hwy), list(average_hwy_year = ~round(mean(.x), 1))))

```

<br>

The second variable is now the fill colour. 

```{r}

ggplot(data=hwy_class_year, aes(x=class, y=average_hwy_year, fill=year)) +
  geom_col(width=0.5, colour="black")

```

<br>

The position argument (setting dodge2) moves the bars to appear alongside each other rather than stacked. 

```{r}

plot <- ggplot(data=hwy_class_year, aes(x=class, y=average_hwy_year, fill=year)) +
  geom_bar(stat="identity", position = "dodge2")

plot

```

<br>

## Adding labels to data 

The geom_text function is what adds the labels: 

 * had to move the aes() bit out of the geom_bar() bit to make this work.  
 * the vjust bit moves it above and below the bar. A lower minus number would be higher above the bar.  
 
```{r}
ggplot(data=hwy_class) +
  aes(x=class, y=average_hwy) +
  geom_bar(stat="identity") + 
  geom_text(aes(label=average_hwy), vjust=-1)
```

<br>

Here the label is below the bar, vjust=1.5. The colour has been switched to white and the size increased also. 

```{r}
ggplot(data=hwy_class) +
  aes(x=class, y=average_hwy) +
  geom_bar(stat="identity") + 
  geom_text(aes(label=average_hwy), vjust=1.5, colour="white",  size=6)
```

#### Adding them to our dodged plot above. 

We also have to 'dodge' the position of the labels within geom_text, again in the position argument. 

```{r}

plot + 
  geom_text(aes(label=average_hwy_year), vjust=1.6, color="white",
            position = position_dodge2(0.9), size=3.5)

```

<br><br>

## Other formatting 

<br> 

### Bar width

Amend bar width within geom_col argument with 'width'. 

The default is 0.9, smaller values make bars smaller. Max = 1 and removes all space between bars. 

If you make the entire graph wider or narrower, the bar dimensions will scale proportionally. 

```{r}

more_space <- ggplot(data=hwy_class) +
  geom_col(mapping=aes(x=class, y=average_hwy), width=0.9)

less_space <- ggplot(data=hwy_class) +
  geom_col(mapping=aes(x=class, y=average_hwy), width=0.5)

ggarrange(more_space, less_space, ncol = 1)

```

<br>

### Change colours   

In the below the colour affects only the outline, without the fill the bars would be dark grey still.  

```{r}
ggplot(data=hwy_class) +
  geom_col(mapping=aes(x=class, y=average_hwy), colour="blue", fill="white", width = 0.5)
```

<br>

Or bar colour can be changed by groups.  

 * Add colour=var, where var is the variable you want to vary the colours by. 
```{r}
ggplot(data=hwy_class) +
  aes(x=class, y=average_hwy, colour=class) +
  geom_col(fill="white", width=0.5) 
```

<br>

### Apply themes

```{r}

p <- ggplot(data=hwy_class) +
  geom_col(mapping=aes(x=class, y=average_hwy), fill="steelblue") 

p + theme_minimal()

```

<br>

## Select categories to show

You can also select which items to show:  

```{r}

ggplot(data=hwy_class) +
  geom_col(mapping=aes(x=class, y=average_hwy, label=average_hwy)) + 
  scale_x_discrete(limits=c("suv", "minivan", "pickup"))

```

<br>

## Alternative statistics

<br>

#### Barplot of counts

Similar to a histogram, the stat="bin" command allows ggplot() to work like table(), but specifically for *numeric* values.  

```{r}

ggplot(data=mpg) +
  aes(x=cyl) +
  geom_bar(stat="bin")

```

<br>

Or with *character* values, try geom_bar without an argument. 

```{r}

ggplot(data=mpg) +
  aes(x=class) +
  geom_bar()

```

<br><br>

## Worked plots

```{r load_packages, warning=FALSE, message=FALSE}
library(haven)
library(labelled)
bsa <- read_sav("/Users/joecrowley/R/Data/BSA Teaching Data/spss/spss24/bsa2017_for_ukda.sav")
# names(bsa)
```

<br><br>

### Bar chart for 1 way crossbreak (or for a frequency table)

```{r}
bsa %>% count(as_factor(leftrigh, levels = "both"))

(leftrigh <- bsa %>% 
  mutate(dolefidl = as_factor(dolefidl), 
         leftrigh = case_when(leftrigh == 9 ~ NA, 
                              TRUE ~ zap_labels(leftrigh))) %>% 
  group_by(dolefidl) %>% 
  summarise(leftrigh = mean(leftrigh, na.rm = TRUE), 
            n = n())) 
```



```{r}
dynamic_str_wrap <- function(text, plot_width = 7, num_labels = 1, char_width = 0.1, margin = 0.1) {
  max_label_width <- (plot_width * (1 - margin)) / num_labels
  max_chars <- max_label_width / char_width
  str_wrap(text, width = floor(max_chars))
}

wrap_titles <- function(title, plot_width = 7, char_width = 0.09) {
  max_chars <- floor(plot_width / char_width)
  wrapped_title <- str_wrap(title, width = max_chars)
  return(wrapped_title)
}

num_labels <- length(levels(fct_drop(leftrigh$dolefidl)))

x_lab <- var_label(bsa[[names(leftrigh[1])]])
x_lab_wrap <- wrap_titles(x_lab)

y_lab <- var_label(bsa[[names(leftrigh[2])]])
y_lab_wrap <- wrap_titles(y_lab, plot_width = 5)

title <- paste0(y_lab, " by ", x_lab)
title_wrap <- wrap_titles(title)

buffer <- mean(leftrigh$leftrigh, na.rm = TRUE) / 4
ylim <- c(0, ceiling(max(leftrigh$leftrigh, na.rm = TRUE) + buffer))

(plot <- leftrigh %>% 
  filter(!is.na(dolefidl)) %>%
  mutate(dolefidl = map_chr(as.character(dolefidl), dynamic_str_wrap, plot_width = 7, num_labels = num_labels, char_width = 0.08, margin = 0.1)) %>%
  mutate(dolefidl = factor(dolefidl, levels = unique(dolefidl))) %>%
  ggplot(aes(x = dolefidl, y = leftrigh)) + 
  geom_col() + 
  geom_text(aes(label = round(leftrigh, 1)), vjust = -1.5, size = 11, size.unit = "pt") + 
  theme_gray() + 
  labs(title = title_wrap, 
       x = x_lab_wrap, 
       y = y_lab_wrap)  + 
  coord_cartesian(ylim = ylim) + 
  theme(
    plot.margin = margin(1,1,1,1),
    text = element_text(size = 11), # This seems to have no effect??? 
    plot.title = element_text(size = 13, face = "bold"),
    axis.title =  element_text(size = 11, face = "bold"), 
    axis.text.x = element_text(margin = margin(b = 10), size = 11), 
    axis.text.y = element_text(margin = margin(l = 10),  size = 11)
  )) 

# Retrospectivly add colour
plot + geom_col(fill = "#619CFF") # ggplot blue

plot + geom_col(fill = "#7082d4") # natcen blue 

plot + geom_col(aes(fill = dolefidl)) + 
  theme(legend.position = "none")

plot + 
  geom_col(aes(fill = dolefidl)) + 
  scale_fill_brewer(palette = "Set2") + 
  theme(legend.position = "none")

```

### Bar chart for 2 way crosstab - clustered

```{r}

(leftrigh_sex <- bsa %>% 
  mutate(dolefidl = as_factor(dolefidl), 
         leftrigh = case_when(leftrigh == 9 ~ NA, 
                              TRUE ~ zap_labels(leftrigh)), 
         sex = as_factor(Rsex)) %>% 
  group_by(sex, dolefidl) %>% 
  summarise(leftrigh = mean(leftrigh, na.rm = TRUE), 
            n = n())) 

num_labels <- length(levels(fct_drop(leftrigh_sex$dolefidl)))

x_lab <- var_label(bsa[[names(leftrigh_sex[2])]])
x_lab_wrap <- wrap_titles(x_lab)

y_lab <- var_label(bsa[[names(leftrigh_sex[3])]])
y_lab_wrap <- wrap_titles(y_lab, plot_width = 5)

title <- paste0(y_lab, " by ", x_lab)
title_wrap <- wrap_titles(title)

buffer <- mean(leftrigh_sex$leftrigh, na.rm = TRUE) / 4
ylim <- c(0, ceiling(max(leftrigh_sex$leftrigh, na.rm = TRUE) + buffer))

(plot <- leftrigh_sex %>% 
  filter(!is.na(dolefidl)) %>%
  mutate(dolefidl = map_chr(as.character(dolefidl), dynamic_str_wrap, plot_width = 7, num_labels = num_labels, char_width = 0.08, margin = 0.1)) %>%
  mutate(dolefidl = factor(dolefidl, levels = unique(dolefidl))) %>%
  ggplot(aes(x = dolefidl, y = leftrigh, fill = sex)) + 
  geom_col(position = "dodge2") + 
  geom_text(aes(label = round(leftrigh, 1)), position = position_dodge2(0.9), vjust = -1, size = 11, size.unit = "pt") + 
  theme_gray() + 
  labs(title = title_wrap, 
       x = x_lab_wrap, 
       y = y_lab_wrap)  + 
  coord_cartesian(ylim = ylim) + 
  theme(
    plot.margin = margin(1,1,1,1),
    text = element_text(size = 11), # This seems to have no effect??? 
    plot.title = element_text(size = 13, face = "bold"),
    axis.title =  element_text(size = 11, face = "bold"), 
    axis.text.x = element_text(margin = margin(b = 10), size = 11), 
    axis.text.y = element_text(margin = margin(l = 10),  size = 11), 
    legend.position = "bottom", 
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 11, face = "bold"),
    legend.margin = unit(0.1,  "cm")
  )) 

# Retrospectivly amend colour

plot +  scale_fill_brewer(palette = "Set2")

plot +  scale_fill_brewer(palette = "Accent")

plot +  scale_fill_brewer(palette = "Pastel1")

plot +  scale_fill_brewer(palette = "Paired")

```

<br>

### Drawing in NatCen colours

```{r}

nc_colours <- data.frame(
  blue =   c("#7082d4", "#8f99de", "#abb5e5", "#c7cced", "#e3e5f5"), 
  pink =   c("#f25c91", "#f57da8", "#f79ebf", "#fabfd4", "#fcdee8"),
  mint =   c("#00ab85", "#33bd9e", "#66ccb5", "#99decf", "#ccede8"), 
  purple = c("#b053a1", "#bf75b5", "#d199c7", "#debad9", "#f0deed"), 
  orange = c("#ff8200", "#ff9c33", "#ffb566", "#ffcc99", "#ffe5cc")) 

nbl <- nc_colours[,1] # extract blue
npi <- nc_colours[,2] 
nmg <- nc_colours[,3] 
npu <- nc_colours[,4] 
nor <- nc_colours[,5] 

nbl_oppo2 <- nc_colours[c(1,4),1] # extract blue
npi_oppo2 <- nc_colours[c(1,4),2] 
nmg_oppo2 <- nc_colours[c(1,4),3] 
npu_oppo2 <- nc_colours[c(1,4),4] 
nor_oppo2 <- nc_colours[c(1,4),5] 

nbl_seq3 <- nc_colours[c(1,3,5),1] # extract blue
npi_seq3 <- nc_colours[c(1,3,5),2] 
nmg_seq3 <- nc_colours[c(1,3,5),3] 
npu_seq3 <- nc_colours[c(1,3,5),4] 
nor_seq3 <- nc_colours[c(1,3,5),5] 

ncl_5to1 <- unlist(c(as.vector(nc_colours[1,]), as.vector(nc_colours[2,]), as.vector(nc_colours[3,]), as.vector(nc_colours[4,]), as.vector(nc_colours[5,]))) %>% unname
ncl_oppo2 <- nc_colours %>% slice(1,4) %>% pivot_longer(cols = everything()) %>% arrange(name) %>% select(-name) %>% as_vector() %>% unname
ncl_seq3 <- nc_colours %>% slice(1,3,5) %>% pivot_longer(cols = everything()) %>% arrange(name) %>% select(-name) %>% as_vector() %>% unname

# Two functions for diverging colours, Natcen... 
scale_fill_ga_diverging <- function(low_color = npi[1], 
                                    medium_color = "white",
                                    high_color = nmg[1]) {
  
  scale_fill_gradient2(low = low_color, 
                       mid = medium_color,
                       high = high_color)
  
}

scale_fill_ga_diverging <- function(low_color = nbl[1,5], 
                                    medium_color = "white",
                                    high_color = nor[1,1]) {
  
  scale_fill_gradient2(low = low_color, 
                       mid = medium_color,
                       high = high_color)
  
}

plot + scale_fill_manual(values = ncl_oppo2[7:8])

plot + scale_fill_manual(values = nbl)

plot + scale_fill_manual(values = nbl_oppo2)

plot + scale_fill_manual(values = ncl_5to1)

```

<br>

### Bar chart for 1 way crosstab - stack

```{r}

dolefidl_x_leftrig2 <- bsa %>% 
  mutate(dolefidl = as_factor(dolefidl), 
         leftrig2 = as_factor(leftrig2),
         sex = as_factor(Rsex)) %>% 
  count(leftrig2, dolefidl) %>%
  filter(!is.na(leftrig2) & !is.na(dolefidl)) %>% 
  group_by(leftrig2) %>% 
  mutate(p = 100*(n / sum(n)), 
         psum = sum(p)) 

dolefidl_x_leftrig2

num_labels <- length(levels(fct_drop(dolefidl_x_leftrig2$leftrig2)))

x_lab <- var_label(bsa[[names(dolefidl_x_leftrig2[1])]])
x_lab_wrap <- wrap_titles(x_lab)

y_lab <- var_label(bsa[[names(dolefidl_x_leftrig2[2])]])
y_lab_wrap <- wrap_titles(y_lab, plot_width = 5)

title <- paste0(y_lab, " by ", x_lab)
title_wrap <- wrap_titles(title)

ylim <- c(0, 100)

dolefidl_x_leftrig2 %>%
  ggplot(aes(x = leftrig2, y = p, fill = dolefidl)) +
  geom_col() + 
  coord_flip()
  
(plot <- dolefidl_x_leftrig2 %>% 
  group_by(leftrig2) %>% 
  mutate(cum_value = cumsum(p),
         lab_pos = (cum_value - p / 2)) %>% # Without reverse of factor order, add 100 - before current formula
  mutate(leftrig2 = map_chr(as.character(leftrig2), dynamic_str_wrap, plot_width = 7, num_labels = num_labels, char_width = 0.08, margin = 0.1)) %>%
  mutate(leftrig2 = factor(leftrig2, levels = unique(leftrig2))) %>%
  ggplot(aes(x = leftrig2, y = p, fill = fct_rev(dolefidl))) + 
  geom_col(position = "stack") + 
  geom_text(aes(label = round(p,0), y = lab_pos), size = 11, size.unit = "pt") + # vjust = -1,
  coord_flip() +
  theme_gray() + 
  labs(title = title_wrap, 
       x = x_lab_wrap, 
       y = y_lab_wrap)  + 
  theme(
    plot.margin = margin(1,1,1,1),
    text = element_text(size = 11), # This seems to have no effect??? 
    plot.title = element_text(size = 13, face = "bold"),
    axis.title =  element_text(size = 11, face = "bold"), 
    axis.text.x = element_text(margin = margin(b = 10), size = 11), 
    axis.text.y = element_text(margin = margin(l = 10),  size = 11), 
    legend.position = "bottom", 
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 11, face = "bold"),
    legend.margin = unit(0.1,  "cm")
  )) 

plot +  scale_fill_brewer(palette = "Set2")
plot +  scale_fill_brewer(palette = "Accent")
plot +  scale_fill_brewer(palette = "Pastel1")
plot + scale_fill_manual(values = rev(c(nbl,"grey")))
plot + scale_fill_manual(values = rev(c(npu,"grey")))
plot + scale_fill_manual(values = rev(c(nmg,"grey")))

```


<br><br>

### Exploration of **str_wrap**

Takes a positive integer giving the target line width (in number of characters).   

A width less than or equal to 1 will put each word on its own line.   

```{r}

leftrigh %>% 
  filter(!is.na(dolefidl)) %>%
  ggplot(aes(x = str_wrap(dolefidl, width = 16), y = leftrigh)) + 
  geom_col() + 
  labs(x = "dolefidl")

# A standard ggplot is 7 inches wide. # An inch is 2.54cm
library(grid)
cats <- levels(fct_drop(leftrigh$dolefidl))
text_grob <- textGrob(cats[2], gp=gpar(fontfamily="Sans", fontsize=11))
width <- grobWidth(text_grob)
height <- grobHeight(text_grob)
width_cm <- convertWidth(width, "cm", valueOnly = TRUE)
height_cm <- convertHeight(height, "cm", valueOnly = TRUE)

cat_lengths <- map_dbl(cats, ~convertWidth(grobWidth(textGrob(.x, gp=gpar(fontfamily="Arial", fontsize=11))),"cm", valueOnly = TRUE))
names(cat_lengths) <- cats	
cat_lengths

sum(cat_lengths) / 2.54

```

<br><br>

## Functions

<br> 

### A frequency bar chart 

Decide to use: age and where someone would go for advice, I can then use for a crosstab later as well. 

```{r}

bsa <- bsa %>% mutate(across(c(RAgecat5,hadvice), ~as_factor(.x)), 
                      hadvice2 = fct_collapse(hadvice, 
                                            `Someone else` = c("Neighbour", "Someone I work with", "Someone else")))

bsa %>% count(hadvice2)
bsa %>% count(RAgecat5)
bsa %>% count(RAgecat5, hadvice)

```

<br>

**Step 1** will be to create a function to generate a table of data, to then plot. 

```{r}

(hadvice <- bsa %>% 
  filter(!is.na(hadvice2)) %>%
  group_by(hadvice2) %>% 
  summarise(n = n()) %>% 
  mutate(p = 100 * (n / sum(n))))

hadvice$x_label <- var_label(bsa$hadvice2)

hadvice

freq <- function(dat, var) { 

x_var <- as.character(substitute(var))

table <- dat %>%
  filter(!is.na({{var}})) %>%
  group_by({{var}}) %>%
  summarise(n = n()) %>%
  mutate(p = 100 * (n / sum(n)))

table$x_label <- var_label(dat[[x_var]])

return(table)
  
}

hadvice <- freq(dat = bsa, var = hadvice2)

hadvice

```

<br> 

**Step 2:** Decant that data into a ggplot of approved type 

```{r}

dynamic_str_wrap <- function(text, plot_width = 7, num_labels = 1, char_width = 0.1, margin = 0.1) {
  max_label_width <- (plot_width * (1 - margin)) / num_labels
  max_chars <- max_label_width / char_width
  str_wrap(text, width = floor(max_chars))
}

wrap_titles <- function(title, plot_width = 7, char_width = 0.09) {
  max_chars <- floor(plot_width / char_width)
  wrapped_title <- str_wrap(title, width = max_chars)
  return(wrapped_title)
}

num_labels <- length(levels(fct_drop(hadvice$hadvice2)))

x_lab <- unique(hadvice[,4])
x_lab_wrap <- wrap_titles(x_lab)

# y_lab <- var_label(bsa[[names(hadvice[2])]])
# y_lab_wrap <- wrap_titles(y_lab, plot_width = 5)

title <- paste0(x_lab)
title_wrap <- wrap_titles(title)

buffer <- mean(hadvice$p, na.rm = TRUE) / 4
ylim <- c(0, ceiling(max(hadvice$p, na.rm = TRUE) + buffer))

(plot <- hadvice %>% 
  filter(!is.na(hadvice2)) %>%
  mutate(hadvice2 = map_chr(as.character(hadvice2), dynamic_str_wrap, plot_width = 7, num_labels = num_labels, char_width = 0.08, margin = 0.1)) %>%
  mutate(hadvice2 = factor(hadvice2, levels = unique(hadvice2))) %>%
  ggplot(aes(x = hadvice2, y = p)) + 
  geom_col() + 
  geom_text(aes(label = round(p, 1)), vjust = -1.5, size = 11, size.unit = "pt") + 
  theme_gray() + 
  labs(title = title_wrap, 
       x = NULL, 
       y = "%"
       )  + 
  coord_cartesian(ylim = ylim) + 
  theme(
    plot.margin = margin(1,1,1,1),
    text = element_text(size = 11), # This seems to have no effect??? 
    plot.title = element_text(size = 13, face = "bold"),
    axis.title =  element_text(size = 11, face = "bold"), 
    axis.title.y =  element_text(angle = 0, vjust = 0.5), 
    axis.text.x = element_text(margin = margin(b = 10), size = 11), 
    axis.text.y = element_text(margin = margin(l = 10),  size = 11)
  )) 

```

** Step 3:** Decant that bar chart code into a function

```{r}

bar <- function(dat, colour = NULL, palette = NULL) { 

# Define some potential custom palettes
nbl <- nc_colours[,1] # extract blue
npi <- nc_colours[,2] 
nmg <- nc_colours[,3] 
npu <- nc_colours[,4] 
nor <- nc_colours[,5] 

nbl_oppo2 <- nc_colours[c(1,4),1] # extract blue
npi_oppo2 <- nc_colours[c(1,4),2] 
nmg_oppo2 <- nc_colours[c(1,4),3] 
npu_oppo2 <- nc_colours[c(1,4),4] 
nor_oppo2 <- nc_colours[c(1,4),5] 

nbl_seq3 <- nc_colours[c(1,3,5),1] # extract blue
npi_seq3 <- nc_colours[c(1,3,5),2] 
nmg_seq3 <- nc_colours[c(1,3,5),3] 
npu_seq3 <- nc_colours[c(1,3,5),4] 
nor_seq3 <- nc_colours[c(1,3,5),5] 

ncl_5to1 <- unlist(c(as.vector(nc_colours[1,]), as.vector(nc_colours[2,]), as.vector(nc_colours[3,]), as.vector(nc_colours[4,]), as.vector(nc_colours[5,]))) %>% unname
ncl_oppo2 <- nc_colours %>% slice(1,4) %>% pivot_longer(cols = everything()) %>% arrange(name) %>% select(-name) %>% as_vector() %>% unname
ncl_seq3 <- nc_colours %>% slice(1,3,5) %>% pivot_longer(cols = everything()) %>% arrange(name) %>% select(-name) %>% as_vector() %>% unname
    
dynamic_str_wrap <- function(text, plot_width = 7, num_labels = 1, char_width = 0.1, margin = 0.1) {
  max_label_width <- (plot_width * (1 - margin)) / num_labels
  max_chars <- max_label_width / char_width
  str_wrap(text, width = floor(max_chars))
}

wrap_titles <- function(title, plot_width = 7, char_width = 0.09) {
  max_chars <- floor(plot_width / char_width)
  wrapped_title <- str_wrap(title, width = max_chars)
  return(wrapped_title)
}

num_labels <- length(levels(fct_drop(hadvice[[1]])))

x_lab <- unique(dat[,4])
x_lab_wrap <- wrap_titles(x_lab)

# y_lab <- var_label(bsa[[names(hadvice[2])]])
# y_lab_wrap <- wrap_titles(y_lab, plot_width = 5)

title <- paste0(x_lab)
title_wrap <- wrap_titles(title)

buffer <- mean(dat$p, na.rm = TRUE) / 4
ylim <- c(0, ceiling(max(dat$p, na.rm = TRUE) + buffer))

x_var <- names(dat)[1]

(plot <- dat %>% 
  filter(!is.na(1)) %>%
  mutate(!!sym(x_var) := map_chr(as.character(!!sym(x_var)), dynamic_str_wrap, plot_width = 7, num_labels = num_labels, char_width = 0.08, margin = 0.1)) %>%
  mutate(!!sym(x_var) := factor(!!sym(x_var), levels = unique(!!sym(x_var)))) %>%
  ggplot(aes(x = !!sym(x_var), y = p)) + 
  geom_col() + 
  geom_text(aes(label = round(p, 1)), vjust = -1.5, size = 11, size.unit = "pt") + 
  theme_gray() + 
  labs(title = title_wrap, 
       x = NULL, 
       y = "%"
       )  + 
  coord_cartesian(ylim = ylim) + 
  theme(
    plot.margin = margin(1,1,1,1),
    text = element_text(size = 11), # This seems to have no effect??? 
    plot.title = element_text(size = 13, face = "bold"),
    axis.title =  element_text(size = 11, face = "bold"), 
    axis.title.y =  element_text(angle = 0, vjust = 0.5), 
    axis.text.x = element_text(margin = margin(b = 10), size = 11), 
    axis.text.y = element_text(margin = margin(l = 10),  size = 11)
  )) 

if (!is_null(colour)) {
  
  if(colour == "pal") {
plot <- plot + 
  geom_col(aes(fill = !!sym(x_var))) + 
  theme(legend.position = "none")

 if(!is_null(palette)) {
   if (palette %in% rownames(RColorBrewer::brewer.pal.info)) {
     
plot <- plot + scale_fill_brewer(palette = palette)

   } else {

plot <- plot + scale_fill_manual(values = get(palette))
     
      }
    }
  } else {
    
plot <- plot + geom_col(fill = colour) 
  }
  
}

return(plot)
  
}

bar(hadvice)
bar(hadvice, colour = "pal")
bar(hadvice, colour = "pal", palette = "Set2")
bar(hadvice, colour = "pal", palette = "Accent")
bar(hadvice, colour = "pal", palette = "ncl_oppo2")
bar(hadvice, colour = "#7082d4")

```

<br> 

**Step 4:** see if it works well with other variables...

```{r}

bsa %>% freq(RAgecat5) %>% 
  bar(colour = "pal")

bsa %>% freq(RAgecat5) %>% bar(colour = "pal", palette = "Set2")
bsa %>% freq(RAgecat5) %>% bar(colour = "pal", palette = "nbl")
bsa %>% freq(RAgecat5) %>% bar(colour = "pal", palette = "npi")
bsa %>% freq(RAgecat5) %>% bar(colour = "pal", palette = "YlGn")
bsa %>% freq(RAgecat5) %>% bar(colour = "pal", palette = "OrRd")

```

<br> 

### X by Y bar chart, labelling Y values

```{r}

# Using the leftrigh table defined above...  
leftrigh

leftrigh <- bsa %>% 
  filter(!is.na(dolefidl)) %>% 
  mutate(dolefidl = fct_drop(as_factor(dolefidl))) %>% 
  group_by(dolefidl) %>%
  summarise(leftrigh = mean(leftrigh, na.rm = TRUE), n = n()) 

leftrigh$x_label <- var_label(bsa$dolefidl)
leftrigh$y_label <- var_label(bsa$leftrigh)

leftrigh

mean_func <- function(dat, x, y) { 
  
  x_var <- as.character(substitute(x))
  y_var <- as.character(substitute(y))

  table <- dat %>% 
  filter(!is.na({{x}})) %>% 
  mutate(!!sym(x_var) := fct_drop(as_factor({{x}}))) %>%
  group_by({{x}}) %>%
  summarise(!!sym(y_var) := mean({{y}}, na.rm = TRUE), n = n())

  table$x_label <- var_label(dat[[x_var]])
  table$y_label <- var_label(dat[[y_var]])

  return(table)
  
}

leftrigh <- mean_func(dat = bsa, x = dolefidl, y = leftrigh)

```

```{r}

plot_xy <- function(table, 
                    .fill = NULL, 
                    .pal_var = NULL) {

# Define some potential custom palettes
nbl <- nc_colours[,1] # extract blue
npi <- nc_colours[,2] 
nmg <- nc_colours[,3] 
npu <- nc_colours[,4] 
nor <- nc_colours[,5] 

nbl_oppo2 <- nc_colours[c(1,4),1] # extract blue
npi_oppo2 <- nc_colours[c(1,4),2] 
nmg_oppo2 <- nc_colours[c(1,4),3] 
npu_oppo2 <- nc_colours[c(1,4),4] 
nor_oppo2 <- nc_colours[c(1,4),5] 

nbl_seq3 <- nc_colours[c(1,3,5),1] # extract blue
npi_seq3 <- nc_colours[c(1,3,5),2] 
nmg_seq3 <- nc_colours[c(1,3,5),3] 
npu_seq3 <- nc_colours[c(1,3,5),4] 
nor_seq3 <- nc_colours[c(1,3,5),5] 

ncl_5to1 <- unlist(c(as.vector(nc_colours[1,]), as.vector(nc_colours[2,]), as.vector(nc_colours[3,]), as.vector(nc_colours[4,]), as.vector(nc_colours[5,]))) %>% unname
ncl_oppo2 <- nc_colours %>% slice(1,4) %>% pivot_longer(cols = everything()) %>% arrange(name) %>% select(-name) %>% as_vector() %>% unname
ncl_seq3 <- nc_colours %>% slice(1,3,5) %>% pivot_longer(cols = everything()) %>% arrange(name) %>% select(-name) %>% as_vector() %>% unname
      
   dynamic_str_wrap <- function(text, plot_width = 7, num_labels = 1, char_width = 0.1, margin = 0.1) {
    max_label_width <- (plot_width * (1 - margin)) / num_labels
    max_chars <- max_label_width / char_width
    str_wrap(text, width = floor(max_chars))
  }
  
  wrap_titles <- function(title, plot_width = 7, char_width = 0.09) {
    max_chars <- floor(plot_width / char_width)
    wrapped_title <- str_wrap(title, width = max_chars)
    return(wrapped_title)
  } 
  
  num_labels <- length(levels(fct_drop(table[[1]])))

  x_lab <- unique(table$x_label)
  x_lab_wrap <- wrap_titles(x_lab)

  y_lab <- unique(table$y_label)
  y_lab_wrap <- wrap_titles(y_lab, plot_width = 5)

  title <- paste0(y_lab, " by ", x_lab)
  title_wrap <- wrap_titles(title)
  
  buffer <- mean(table[[2]], na.rm = TRUE) / 4
  ylim <- c(0, ceiling(max(table[[2]], na.rm = TRUE) + buffer))
  
  x_var <- names(table)[1]
  y_var <- names(table)[2]
  
(plot <- table %>% 
  filter(!is.na(!!sym(x_var))) %>%
  mutate(!!sym(x_var) := map_chr(as.character(!!sym(x_var)), dynamic_str_wrap, plot_width = 7, num_labels = num_labels, char_width = 0.08, margin = 0.1)) %>%
  mutate(!!sym(x_var) := factor(!!sym(x_var), levels = unique(!!sym(x_var)))) %>%
  ggplot(aes(x = !!sym(x_var), y = !!sym(y_var))) + 
  geom_col() + 
  geom_text(aes(label = round(!!sym(y_var), 1)), vjust = -1.5, size = 11, size.unit = "pt") + 
  theme_gray() + 
  labs(title = title_wrap, 
       x = x_lab_wrap, 
       y = y_lab_wrap)  + 
  coord_cartesian(ylim = ylim) + 
  theme(
    plot.margin = margin(1,1,1,1),
    text = element_text(size = 11), # This seems to have no effect??? 
    plot.title = element_text(size = 13, face = "bold"),
    axis.title =  element_text(size = 11, face = "bold"), 
    axis.text.x = element_text(margin = margin(b = 10), size = 11), 
    axis.text.y = element_text(margin = margin(l = 10),  size = 11)
  )) 
  
  if (!is.null(.fill)) { 
    
  plot <- plot + geom_col(fill = .fill) # ggplot blue
    
  }
  
  if (!is.null(.pal_var)) { 
    
  plot <- plot + geom_col(aes(fill = !!sym(x_var))) + 
    theme(legend.position = "none")  
  
    if(is.character(.pal_var)) { 
   
           if (.pal_var %in% rownames(RColorBrewer::brewer.pal.info)) {
   plot <- plot + scale_fill_brewer(palette = .pal_var)   
           } else {
      
plot <- plot + scale_fill_manual(values = get(.pal_var))
     
      }
   
    }
      
    }
  
  return(list(num_labels, x_lab_wrap, y_lab_wrap, title_wrap, buffer, ylim, plot))
  
}

plot_xy(leftrigh)
plot_xy(leftrigh, .fill = "#619CFF")
plot_xy(leftrigh, .pal_var = TRUE)
plot_xy(leftrigh, .pal_var = "Set2")
plot_xy(leftrigh, .pal_var = "ncl_oppo2")


```



