---
title: "Theme: what it controls, and the element_ functions"
format: html
editor: visual
---

`theme()` controls every piece of non-data ink on a plot — anything not determined by an aesthetic mapping. It can't touch scales, coordinates, or geoms; it's purely visual. Every argument takes one of four element functions.

```{r setup}
#| message: false
library(tidyverse)
```

### The areas theme() works on

**Text.** Every piece of writing on the plot — title, axis labels, legend labels, facet strip labels — can be styled independently, though they all fall back to one shared default if left untouched. e.g. `axis.text = element_text(size = 8)` shrinks just the tick labels, leaving the title and legend alone.

**Axis details.** The physical marks that frame an axis — its tick marks and the line running along it — as distinct from the text sitting next to them. e.g. `axis.ticks = element_blank()` removes the tick marks entirely while the axis text stays put.

**Panel.** The plotting area itself: its background, border, and the gridlines sitting behind the data — the canvas the geoms are drawn on top of, not the data. e.g. `panel.grid.minor = element_blank()` drops the finer gridlines, leaving only the major ones for orientation.

**Legend.** Where the legend sits and how its internal pieces — the coloured swatches, their spacing, their labels — are laid out, entirely separate from *what* the legend shows (that's controlled by `scale_*()`/`labs()`, not `theme()`). e.g. `legend.position = "bottom"` moves the whole legend under the plot instead of to the side.

**Facet strips.** The small labelled boxes sitting above each facet panel, identifying which subset of the data that panel represents. e.g. `strip.background = element_rect(fill = "grey85")` gives each strip a shaded backing so it stands out from the panel below it.

**Plot-level.** The outer canvas the entire plot sits on — background and the breathing room around its edges — outside the panel itself. e.g. `plot.margin = margin(10, 10, 10, 10)` adds even padding around the whole thing.

A couple of mechanics worth knowing beyond the individual areas: elements inherit down a hierarchy (setting `text` cascades to `axis.text`, `legend.text`, etc. unless a more specific one overrides it), `theme_set()` changes the default theme for the rest of a session instead of adding `+ theme_minimal()` everywhere, and `%+replace%` fully replaces an inherited element rather than merging into it — relevant if building a custom theme rather than tweaking a built-in one.

### The element_ functions

**`element_text()`** — anything rendered as text. Arguments: `family` (font), `face` (`"plain"`/`"bold"`/`"italic"`/`"bold.italic"`), `colour`, `size` (points), `hjust`/`vjust` (0–1 justification), `angle` (degrees), `lineheight`, `margin` (spacing around the text itself, via `margin()`).

**`element_line()`** — anything drawn as a line: axis lines, tick marks, gridlines. Arguments: `colour`, `linewidth` (renamed from `size` in ggplot2 3.4+), `linetype` (`"solid"`/`"dashed"`/`"dotted"`/etc.), `lineend` (`"round"`/`"butt"`/`"square"`), `arrow`.

**`element_rect()`** — anything filled as a rectangle: panel background, plot background, legend background, facet strip background. Arguments: `fill`, `colour` (border), `linewidth` (border thickness), `linetype`.

**`element_blank()`** — draws nothing, and takes no arguments. Removing an element this way also frees the space it would have occupied — the difference from setting `colour = "transparent"` or `fill = NA`, which keep the element's footprint.

### Seeing all four at once

One deliberately loud example — not a style to actually use, but each change is exaggerated enough that it can only be attributed to one element function.

```{r}
ggplot(mpg, aes(x = class)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Vehicles by class") +
  theme(
    # element_text() — anything rendered as text
    plot.title = element_text(size = 20, face = "bold.italic", colour = "firebrick"),

    # element_line() — anything drawn as a line
    panel.grid.major = element_line(colour = "darkorange", linewidth = 1.2, linetype = "dashed"),

    # element_rect() — anything filled as a rectangle
    panel.background = element_rect(fill = "lightyellow", colour = "black", linewidth = 1.5),

    # element_blank() — removes the element and its space entirely
    axis.ticks = element_blank()
  )
```

The title screams "text," the dashed orange gridlines are obviously a line element, the yellow bordered panel is obviously a rectangle, and the vanished tick marks — not just invisible, genuinely gone, no reserved space — show what `element_blank()` does that a transparent colour wouldn't.

### One example per functional area

That plot showed *how* to change things (which element function). This section shows *where* — one deliberate change per functional area from the section above, each cranked up loud enough to be unmissable, before combining all six at the end.

```{r}
base_plot <- ggplot(mpg, aes(x = class, fill = drv)) +
  geom_bar() +
  facet_wrap(~ year) +
  labs(title = "Vehicles by class and drivetrain, by year")

base_plot
```

**Text.** The title, blown up and recoloured well past anything you'd actually ship — impossible to miss, and nothing else on the plot moves.

```{r}
base_plot +
  theme(plot.title = element_text(size = 28, face = "bold.italic", colour = "deeppink"))
```

**Axis details.** Tick marks turned into thick red spikes — the axis text next to them is completely untouched, proving this is a separate area from Text.

```{r}
base_plot +
  theme(axis.ticks = element_line(colour = "red", linewidth = 4))
```

**Panel.** The plotting area itself painted bright yellow — the canvas the bars sit on, not the bars.

```{r}
base_plot +
  theme(panel.background = element_rect(fill = "yellow"))
```

**Legend.** The swatches blown up to nearly the size of the plot itself.

```{r}
base_plot +
  theme(legend.key.size = unit(2.5, "cm"))
```

**Facet strips.** The panel-label boxes ("1999"/"2008") lit up bright orange.

```{r}
base_plot +
  theme(strip.background = element_rect(fill = "orange"))
```

**Plot-level.** The entire outer canvas — everything outside the panel — turned hot pink.

```{r}
base_plot +
  theme(plot.background = element_rect(fill = "hotpink"))
```

### Putting it together

All six changes on the one chart at once — deliberately garish, but that's what makes it unambiguous which area each one belongs to.

```{r}
base_plot +
  theme(
    # Text
    plot.title = element_text(size = 28, face = "bold.italic", colour = "deeppink"),

    # Axis details
    axis.ticks = element_line(colour = "red", linewidth = 4),

    # Panel
    panel.background = element_rect(fill = "yellow"),

    # Legend
    legend.key.size = unit(2.5, "cm"),

    # Facet strips
    strip.background = element_rect(fill = "orange"),

    # Plot-level
    plot.background = element_rect(fill = "hotpink")
  )
```
