---
title: "Bar chart functions"
format: html
editor: visual
---

```{r, message=F, warning=FALSE, collapse=T}
library(tidyverse)
library(srvyr)
library(flextable)
library(haven)
library(labelled)
library(huxtable)
library(here)

# Load data
# df2    <- readRDS(here::here("data", "synthetic_multicode_data.rds"))
df2 <- readRDS(here::here("data", "synthetic_bsa_clean.rds"))
df_svy <- df2 %>% as_survey_design(weights = WtFactor)

source(here::here("Scripts/colours.R"))


## Load functions from Descriptives project on GitHub
scripts <- jsonlite::fromJSON(
  "https://api.github.com/repos/Joeacrowley/Descriptives/contents/Scripts"
)
r_scripts <- scripts$download_url[grepl("\\.R$", scripts$name, ignore.case = TRUE)]
quietly(lapply(r_scripts, source))
```

## Philosophy

To arrive at a simple, usable piece of code, that is useful, as early as possible.

To write generic functions - not too bespoke to a given use-case.

Here, I initially wrote something tightly linked to the output format of calc_stats. It depended on column names and content more than it should have. I have now adjusted to make `bar()` more generic, taking any factor column as it's x-axis and any numeric column as the y-axis, with labels simply fed as strings.

This can then be linked quite simply to the output of calc_stats, but is also useful on its own elsewhere.

### Purpose

Take a dataframe with one categorical (ideally factor) variable to use as the x-axis and a second numeric variable to use as the y-axis and presents this nicely in a ggplot. Add variable labels as x- and y- axis titles, and allow a user defined plot title.

These plot values in a dataframe, they don't calculate new statistics from a larger dataframe.

```{r}

(data_frame <- calc_stats(
  data = df2, 
  outcome ="skipmeal2", 
  statistics = "perc"
))

```

## Functions

```{r}
source(here::here("Scripts/bar_charts.R"))
```

<br>

## Examples

```{r}

data_frame %>% glimpse

```

::::: panel-tabset
### Basic

```{r}

bar(dat = data_frame %>% mutate(estimate = estimate * 100), 
    x = o_cat, 
    y = estimate, 
    fill = ncolours$blue2,
    x_lab = data_frame$o_lab %>% unique, 
    y_lab = "%", 
    title = NULL)

```

### Orange

```{r}

bar(dat = data_frame %>% mutate(estimate = estimate * 100), 
    x = o_cat, 
    y = estimate, 
    fill = ncolours$orange1,
    x_lab = data_frame$o_lab %>% unique, 
    y_lab = "Percentage", 
    title = "Plot of sorts")

```

### y-axis label

```{r}

bar(dat = data_frame %>% mutate(estimate = estimate * 100), 
    x = o_cat, 
    y = estimate, 
    x_lab = data_frame$o_lab %>% unique, 
    y_lab = "Percentage - very long label text, arbitrarily added to test how this functions.")

```

### Lib-auth scale

```{r}

(data_frame2 <- calc_stats(
  data = df2, 
  outcome ="libauth", 
  predictors = "RSex",
  statistics = "mean"
  )
 )

data_frame2 %>% glimpse

bar(dat = data_frame2, 
    x = p_cat1, 
    y = estimate, 
    x_lab = "Sex", 
    y_lab = "Mean", 
    title = "Libertarian authoritarian scale") 


```

### Clustered bar

```{r}
glimpse(df2)

data_frame3 <- calc_stats(
  data = df2, outcomes = "welfare3", predictors = c("RSex"), statistics = "perc"
  ) %>% 
  mutate(estimate = round(estimate * 100,1))

data_frame3

bar_clustered(
  dat = data_frame3, 
  x = "o_cat",
  fill = "p_cat1",
  y = "estimate", 
  x_lab = data_frame3$o_lab %>% unique, 
  y_lab = "%", 
  fill_lab =  data_frame3$p_lab1 %>% unique %>% grep("Total", ., value = T, invert = T), 
  title = "Left right scale, crosstabulated by sex"
)

```

<br>

### Other palettes

::: panel-tabset
```{r, results='asis'}

palettes <- list(
  colours_pairs = ncolours$colours_pairs,
  colours10     = ncolours$colours10,
  blues         = ncolours$blues,
  mints         = ncolours$mints,
  oranges       = ncolours$oranges
)

for (pal_name in names(palettes)) {

  cat("####", pal_name, "\n\n")

  bar_clustered(
    dat = data_frame3,
    x = "o_cat",
    fill = "p_cat1",
    y = "estimate",
    x_lab = data_frame3$o_lab %>% unique,
    y_lab = "%",
    fill_lab =  data_frame3$p_lab1 %>% unique %>% grep("Total", ., value = T, invert = T),
    title = "Left right scale, crosstabulated by sex",
    palette = palettes[[pal_name]]
  ) %>% print()

  cat("\n\n")

}
```
:::

<br>

### Stacked plots

::: panel-tabset
#### Legend bottom

```{r}

bar_stacked(
  dat             = data_frame3,
  x               = "p_cat1",
  fill            = "o_cat",
  y               = "estimate",
  x_lab           = data_frame3$p_lab1 %>% unique %>% grep("Total", ., value = T, invert = T),
  y_lab           = "%",
  fill_lab        = data_frame3$o_lab %>% unique,
  title           = "Left right scale, crosstabulated by sex",
  title_size      = 14,
  legend_position = "bottom"
)

```

#### Legend right

Best for long response labels — each label gets the full legend column width.

```{r}

bar_stacked(
  dat             = data_frame3,
  x               = "p_cat1",
  fill            = "o_cat",
  y               = "estimate",
  x_lab           = data_frame3$p_lab1 %>% unique %>% grep("Total", ., value = T, invert = T),
  y_lab           = "%",
  fill_lab        = NULL,
  title           = "Left right scale, crosstabulated by sex",
  title_size      = 14,
  legend_position = "right"
)

```
:::
:::::
