---
title: "Bar chart functions"
format: html
editor: visual
---

```{r, message=F, warning=FALSE, collapse=T}
library(tidyverse)
library(srvyr)
library(flextable)
library(haven)
library(labelled)
library(huxtable)

# Set path to and Load data
source("Set file paths.R")
## Load functions
source("/Users/joecrowley/R/Descriptives/Scripts/source_folder_r.R")
source_folder_r("/Users/joecrowley/R/Descriptives/Scripts", verbose = T)
```

## Philosophy

To arrive at a simple, usable piece of code, that is useful, as early as possible.

To write generic functions - not too bespoke to a given use-case.

Here, I initially wrote something tightly linked to the output format of calc_stats. It depended on column names and content more than it should have. I have now adjusted to make `bar()` more generic, taking any factor column as it's x-axis and any numeric column as the y-axis, with labels simply fed as strings.

This can then be linked quite simply to the output of calc_stats, but is also useful on its own elsewhere.

### Purpose

Take a dataframe with one categorical (ideally factor) variable to use as the x-axis and a second numeric variable to use as the y-axis and presents this nicely in a ggplot. Add variable labels as x- and y- axis titles, and allow a user defined plot title.

These plot values in a dataframe, they don't calculate new statistics from a larger dataframe.

```{r}

(data_frame <- calc_stats(
  data = df2, 
  outcome ="skipmeal2", 
  statistics = "perc"
))

```

## Functions

```{r}
# wrap_text() -----------------------------------------------------------------
# Wraps a string to fit within a ggplot plot area.
# plot_width: estimated plot width in inches
# num_labels: number of bars sharing the axis — defaults to 1, which makes
#             this suitable for titles and axis labels as well as tick labels
# char_width: approximate width of one character in inches — adjust for font size
# margin:     proportion of plot width to treat as unusable

wrap_text <- function(text, plot_width = 7, num_labels = 1, char_width = 0.09, margin = 0.1) {
  usable_width <- plot_width * (1 - margin)
  max_chars    <- floor((usable_width / num_labels) / char_width)
  str_wrap(text, width = max_chars)
}


# resolve_y_lab() -------------------------------------------------------------
# Resolves y-axis label text and theme element based on label length.
# Returns a named list: y_lab_final and y_axis_theme.
# Short labels (<=3 chars) are horizontal, longer labels are vertical and wrapped.

resolve_y_lab <- function(y_lab, plot_height = 4) {
  if (!is.null(y_lab) && nchar(y_lab) > 3) {
    list(
      y_lab_final  = wrap_text(y_lab, plot_width = plot_height),
      y_axis_theme = element_text(angle = 90, hjust = 0.5, size = 11, face = "bold")
    )
  } else if (!is.null(y_lab)) {
    list(
      y_lab_final  = y_lab,
      y_axis_theme = element_text(angle = 0, vjust = 0.5, size = 11, face = "bold")
    )
  } else {
    list(
      y_lab_final  = NULL,
      y_axis_theme = element_text(angle = 0, vjust = 0.5, size = 11, face = "bold")
    )
  }
}


# bar_width_from_n() ----------------------------------------------------------
# Scales bar width to the number of x-axis labels.
# Fewer bars get narrower widths to avoid them looking too fat.

bar_width_from_n <- function(num_labels) {
  case_when(
    num_labels <= 3 ~ 0.4,
    num_labels <= 5 ~ 0.6,
    num_labels <= 8 ~ 0.75,
    TRUE            ~ 0.9
  )
}


# bar() -----------------------------------------------------------------------
# Bar chart for a single categorical x numeric variable.
# x:           unquoted name of the categorical column (ideally a factor) for the x-axis
# y:           unquoted name of the numeric column for the y-axis
# x_lab:       optional character string for the x-axis title
# y_lab:       optional character string for the y-axis title — if longer than 3
#              characters the label is rotated vertical and wrapped to plot height
# title:       optional — left empty if not supplied
# fill:        bar colour — "blue" (default), "softblue", "green", "orange"
# plot_width:  estimated plot width in inches — used for label wrapping, default 7
# plot_height: estimated plot height in inches — used for y-axis label wrapping
#              when the label is long, default 4

bar <- function(dat, x, y, x_lab = NULL, y_lab = NULL, title = NULL,
                fill = "blue", plot_width = 7, plot_height = 4) {

  fill_options <- c(
    blue     = "#2E75B6",
    softblue = "#5B9BD5",
    green    = "#70AD47",
    orange   = "#ED7D31"
  )

  fill <- match.arg(fill, names(fill_options))
  fill_colour <- fill_options[[fill]]

  x_var <- as.character(substitute(x))
  y_var <- as.character(substitute(y))

  x_lab_wrap <- if (!is.null(x_lab)) wrap_text(x_lab, plot_width = plot_width) else NULL
  title_wrap <- if (!is.null(title)) wrap_text(title, plot_width = plot_width) else NULL

  y_lab_resolved <- resolve_y_lab(y_lab, plot_height)
  y_lab_final    <- y_lab_resolved$y_lab_final
  y_axis_theme   <- y_lab_resolved$y_axis_theme

  num_labels <- length(levels(fct_drop(as.factor(dat[[x_var]]))))
  bar_width  <- bar_width_from_n(num_labels)

  buffer <- mean(dat[[y_var]], na.rm = TRUE) / 4
  ylim   <- c(0, ceiling(max(dat[[y_var]], na.rm = TRUE) + buffer))

  dat %>%
    filter(!is.na(!!sym(x_var))) %>%
    mutate(
      !!sym(x_var) := map_chr(
        as.character(!!sym(x_var)),
        wrap_text,
        num_labels = num_labels,
        char_width = 0.08
      ),
      !!sym(x_var) := factor(!!sym(x_var), levels = unique(!!sym(x_var)))
    ) %>%
    ggplot(aes(x = !!sym(x_var), y = !!sym(y_var))) +
    geom_col(fill = fill_colour, width = bar_width) +
    geom_text(aes(label = round(!!sym(y_var), 1)), vjust = -1.5, size = 11, size.unit = "pt") +
    theme_linedraw() +
    labs(
      title = title_wrap,
      x     = x_lab_wrap,
      y     = y_lab_final
    ) +
    coord_cartesian(ylim = ylim) +
    theme(
      plot.margin  = margin(1, 1, 1, 1),
      plot.title   = element_text(size = 13, face = "bold"),
      axis.title   = element_text(size = 11, face = "bold"),
      axis.title.y = y_axis_theme,
      axis.text.x  = element_text(margin = margin(b = 10), size = 11),
      axis.text.y  = element_text(margin = margin(l = 10), size = 11)
    )

}


# bar_clustered() -------------------------------------------------------------
# Clustered bar chart for a categorical x numeric variable, grouped by a third
# factor variable displayed as fill colour.
# x:           unquoted name of the categorical column for the x-axis
# y:           unquoted name of the numeric column for the y-axis
# fill:        unquoted name of the grouping variable mapped to fill colour
# x_lab:       optional character string for the x-axis title
# y_lab:       optional character string for the y-axis title
# fill_lab:    optional character string for the legend title
# title:       optional — left empty if not supplied
# palette:     RColorBrewer palette name for fill colours, default "Set2"
# plot_width:  estimated plot width in inches — used for label wrapping, default 7
# plot_height: estimated plot height in inches — used for y-axis label wrapping
#              when the label is long, default 4

bar_clustered <- function(dat, x, y, fill, x_lab = NULL, y_lab = NULL,
                          fill_lab = NULL, title = NULL,
                          palette = c("Set2"), plot_width = 7, plot_height = 4) {

  x_var    <- as.character(substitute(x))
  y_var    <- as.character(substitute(y))
  fill_var <- as.character(substitute(fill))

  x_lab_wrap <- if (!is.null(x_lab)) wrap_text(x_lab, plot_width = plot_width) else NULL
  title_wrap <- if (!is.null(title)) wrap_text(title, plot_width = plot_width) else NULL

  y_lab_resolved <- resolve_y_lab(y_lab, plot_height)
  y_lab_final    <- y_lab_resolved$y_lab_final
  y_axis_theme   <- y_lab_resolved$y_axis_theme

  num_labels <- length(levels(fct_drop(as.factor(dat[[x_var]]))))

  buffer <- mean(dat[[y_var]], na.rm = TRUE) / 4
  ylim   <- c(0, ceiling(max(dat[[y_var]], na.rm = TRUE) + buffer))

  dat %>%
    filter(!is.na(!!sym(x_var))) %>%
    mutate(
      !!sym(x_var) := map_chr(
        as.character(!!sym(x_var)),
        wrap_text,
        num_labels = num_labels,
        char_width = 0.08
      ),
      !!sym(x_var) := factor(!!sym(x_var), levels = unique(!!sym(x_var))),
      !!sym(fill_var) := as.factor(!!sym(fill_var))
    ) %>%
    ggplot(aes(x = !!sym(x_var), y = !!sym(y_var), fill = !!sym(fill_var))) +
    geom_col(position = "dodge2") +
    geom_text(
      aes(label = round(!!sym(y_var), 1)),
      position = position_dodge2(0.9),
      vjust = -1.5,
      size = 11,
      size.unit = "pt"
    ) +
    scale_fill_brewer(palette = palette) +
    theme_linedraw() +
    labs(
      title = title_wrap,
      x     = x_lab_wrap,
      y     = y_lab_final,
      fill  = fill_lab
    ) +
    coord_cartesian(ylim = ylim) +
    theme(
      plot.margin  = margin(1, 1, 1, 1),
      plot.title   = element_text(size = 13, face = "bold"),
      axis.title   = element_text(size = 11, face = "bold"),
      axis.title.y = y_axis_theme,
      axis.text.x  = element_text(margin = margin(b = 10), size = 11),
      axis.text.y  = element_text(margin = margin(l = 10), size = 11),
      legend.title = element_text(size = 11, face = "bold"),
      legend.position = "bottom"
    )

}

# bar_stacked() ---------------------------------------------------------------
# Stacked horizontal bar chart for a categorical x numeric variable,
# grouped by a third factor variable displayed as fill colour.
# x:           unquoted name of the categorical column for the x-axis
# y:           unquoted name of the numeric column for the y-axis
# fill:        unquoted name of the grouping variable mapped to fill colour
# x_lab:       optional character string for the x-axis title
# y_lab:       optional character string for the y-axis title
# fill_lab:    optional character string for the legend title
# title:       optional — left empty if not supplied
# palette:     RColorBrewer palette name for fill colours, default "Set2"
# plot_width:  estimated plot width in inches — used for label wrapping, default 7
# plot_height: estimated plot height in inches — used for y-axis label wrapping,
#              default 4

bar_stacked <- function(dat, x, y, fill, x_lab = NULL, y_lab = NULL,
                        fill_lab = NULL, title = NULL,
                        palette = "Set2", plot_width = 7, plot_height = 4) {

  x_var    <- as.character(substitute(x))
  y_var    <- as.character(substitute(y))
  fill_var <- as.character(substitute(fill))

  x_lab_wrap <- if (!is.null(x_lab)) wrap_text(x_lab, plot_width = plot_width * 0.2) else NULL
  # x_lab_wrap <- if (!is.null(x_lab)) wrap_text(x_lab, plot_width = plot_width) else NULL
  y_lab_wrap <- if (!is.null(y_lab)) wrap_text(y_lab, plot_width = plot_width) else NULL
  title_wrap <- if (!is.null(title)) wrap_text(title, plot_width = plot_width) else NULL

  num_labels <- length(levels(fct_drop(as.factor(dat[[x_var]]))))
  bar_width  <- bar_width_from_n(num_labels)

  buffer <- mean(dat[[y_var]], na.rm = TRUE) / 4
  ylim   <- c(0, ceiling(max(dat[[y_var]], na.rm = TRUE) + buffer))

  dat %>%
    filter(!is.na(!!sym(x_var))) %>%
    mutate(
      !!sym(x_var) := map_chr(
        as.character(!!sym(x_var)),
        wrap_text,
        num_labels = num_labels,
        char_width = 0.08
      ),
      !!sym(x_var) := factor(!!sym(x_var), levels = unique(!!sym(x_var))),
      !!sym(fill_var) := as.factor(!!sym(fill_var))
    ) %>%
    ggplot(aes(x = !!sym(x_var), y = !!sym(y_var), fill = !!sym(fill_var))) +
    geom_col(position = "stack", width = bar_width) +
    geom_text(
      aes(label = round(!!sym(y_var), 1)),
      position = position_stack(vjust = 0.5),
      size = 11,
      size.unit = "pt"
    ) +
    scale_fill_brewer(palette = palette) +
    coord_flip() +
    theme_gray() +
    labs(
      title = title_wrap,
      x     = x_lab_wrap,
      y     = y_lab_wrap,
      fill  = fill_lab
    ) +
    theme(
      plot.margin  = margin(1, 1, 1, 1),
      plot.title   = element_text(size = 13, face = "bold"),
      axis.title   = element_text(size = 11, face = "bold"),
      axis.title.y = element_text(angle = 0, vjust = 0.5, size = 11, face = "bold"),
      axis.text.x  = element_text(margin = margin(b = 10), size = 11),
      axis.text.y  = element_text(margin = margin(l = 10), size = 11),
      legend.title     = element_text(size = 11, face = "bold"),
      legend.position  = "bottom",
      legend.direction = "horizontal"
    )

}

```

<br>

## Examples

```{r}

data_frame %>% glimpse

```

:::: panel-tabset
### Basic

```{r}

bar(dat = data_frame %>% mutate(estimate = estimate * 100), 
    x = o_cat, 
    y = estimate, 
    fill = "softblue",
    x_lab = data_frame$o_lab %>% unique, 
    y_lab = "%", 
    title = NULL)

```

### Orange

```{r}

bar(dat = data_frame %>% mutate(estimate = estimate * 100), 
    x = o_cat, 
    y = estimate, 
    fill = "orange",
    x_lab = data_frame$o_lab %>% unique, 
    y_lab = "Percentage", 
    title = "Plot of sorts")

```

### y-axis label

```{r}

bar(dat = data_frame %>% mutate(estimate = estimate * 100), 
    x = o_cat, 
    y = estimate, 
    x_lab = data_frame$o_lab %>% unique, 
    y_lab = "Percentage - very long label text, arbitrarily added to test how this functions.")

```

### Lib-auth scale

```{r}

(data_frame2 <- calc_stats(
  data = df2, 
  outcome ="libauth", 
  predictors = "RSex",
  statistics = "mean"
  )
 )

data_frame2 %>% glimpse

bar(dat = data_frame2, 
    x = p_cat1, 
    y = estimate, 
    x_lab = "Sex", 
    y_lab = "Mean", 
    title = "Libertarian authoritarian scale") 


```

### Clustered bar

```{r}
glimpse(df2)

data_frame3 <- calc_stats(
  data = df2, outcomes = "welfare3", predictors = c("RSex"), statistics = "perc"
  ) %>% 
  mutate(estimate = round(estimate * 100,1))

data_frame3

bar_clustered(
  dat = data_frame3, 
  x = "o_cat",
  fill = "p_cat1",
  y = "estimate", 
  x_lab = data_frame3$o_lab %>% unique, 
  y_lab = "%", 
  fill_lab =  data_frame3$p_lab1 %>% unique %>% grep("Total", ., value = T, invert = T), 
  title = "Left right scale, crosstabulated by sex"
)

```

<br>

### Other palettes

Not sure about any of these really.

::: panel-tabset
```{r, results='asis'}

palettes <- c("Set2", "Set1", "Dark2", "Paired", "RdBu", "PuOr", "RdYlGn", "Blues", "YlOrRd")

for (pal in palettes) {
  
  cat("####", pal, "\n\n")
  
  RColorBrewer::display.brewer.pal(n = 8, name = pal)
  title(pal)
  
  cat("\n\n")
  
  bar_clustered(
    dat = data_frame3, 
    x = "o_cat",
    fill = "p_cat1",
    y = "estimate", 
    x_lab = data_frame3$o_lab %>% unique, 
    y_lab = "%", 
    fill_lab =  data_frame3$p_lab1 %>% unique %>% grep("Total", ., value = T, invert = T), 
    title = "Left right scale, crosstabulated by sex", 
    palette = pal
  ) %>% print()
  
    cat("\n\n")
  
}
```
:::

<br>

### Stacked plots

<br>

```{r}

bar_stacked

bar_stacked(
  dat = data_frame3, 
  x = "p_cat1",
  fill = "o_cat",
  y = "estimate", 
  x_lab = data_frame3$p_lab1 %>% unique %>% grep("Total", ., value = T, invert = T),
  y_lab = "%", 
  fill_lab = data_frame3$o_lab %>% unique, 
  title = "Left right scale, crosstabulated by sex", 
  palette = "Set2"
)



```
::::
