---
title: "Outcomes by predictor — worked example"
format: html
editor: visual
---

A worked example looping through a set of categorical and numeric outcome variables, crosstabulated by a single predictor, using the repo bar chart functions and BSA synthetic data. Mirrors the structure of a typical survey analysis output script.

```{r setup, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(srvyr)
library(haven)
library(labelled)
library(here)
library(officer)

# Load data
df2    <- readRDS(here::here("data", "synthetic_bsa_clean.rds"))
df_svy <- df2 %>% as_survey_design(weights = WtFactor)

# Load local scripts
source(here::here("Scripts/colours.R"))
source(here::here("Scripts/bar_charts.R"))

# Load calc_stats and dependencies from GitHub
scripts   <- jsonlite::fromJSON(
  "https://api.github.com/repos/Joeacrowley/Descriptives/contents/Scripts"
)
r_scripts <- scripts$download_url[grepl("\\.R$", scripts$name, ignore.case = TRUE)]
invisible(lapply(r_scripts, source))
```

<br>

## Config

Define the outcome variables and predictor to loop over.

```{r config}

# Categorical outcomes — weighted proportions, displayed as stacked bar
outcomes_cat <- c("welfare3", "skipmeal2", "MuchPov2")

# Numeric outcomes — weighted median, displayed as simple bar
outcomes_num <- c("libauth", "welfare2", "leftrigh")

# Crosstab predictor
crosstab <- "RSex"

# Base font size for PowerPoint export
pptx_base_size <- 12

# Legend position for stacked bar charts — "right" handles long response labels better
pptx_legend_position <- "right"
```

<br>

## Helper functions

Two thin wrappers that handle aggregation via `calc_stats` and pass the result to the appropriate bar chart function, plus a PowerPoint export function.

```{r helpers}

# get_base_string() -----------------------------------------------------------
# Extracts a formatted unweighted base size string from calc_stats output.
# Expects columns p_cat1 and base. Returns e.g. "Male: n=412 | Female: n=468"

get_base_string <- function(dat) {
  dat %>%
    distinct(p_cat1, base) %>%
    mutate(label = paste0(p_cat1, ": n=", base)) %>%
    pull(label) %>%
    paste(collapse = " | ")
}


# export_plots_to_pptx() ------------------------------------------------------
# Saves a list of ggplots to a PowerPoint file, one plot per slide.
# plot_list:    list of ggplot objects
# base_sizes:   optional list of base size strings, one per plot
# filename:     output path for the .pptx file
# slide_width:  slide width in inches, default 10
# slide_height: slide height in inches, default 7.5
# dpi:          image resolution, default 300

export_plots_to_pptx <- function(plot_list, base_sizes = NULL,
                                 filename = "output.pptx",
                                 slide_width = 10, slide_height = 7.5,
                                 dpi = 300) {

  prs       <- read_pptx()
  tmp_files <- c()

  img_height <- if (!is.null(base_sizes)) slide_height - 1.25 else slide_height - 1

  for (i in seq_along(plot_list)) {

    tmp <- tempfile(fileext = ".png")
    ggsave(tmp, plot = plot_list[[i]], width = slide_width - 1,
           height = img_height, dpi = dpi, bg = "white")
    tmp_files <- c(tmp_files, tmp)

    prs <- prs |>
      add_slide(layout = "Blank", master = "Office Theme") |>
      ph_with(
        value    = external_img(tmp, width = slide_width - 1, height = img_height),
        location = ph_location(left = 0.5, top = 0.5,
                               width = slide_width - 1, height = img_height)
      )

    if (!is.null(base_sizes) && !is.na(base_sizes[[i]])) {
      prs <- prs |>
        ph_with(
          value = fpar(
            ftext("Unweighted sample sizes: ",
                  fp_text(font.size = 12, color = "#666666", bold = TRUE)),
            ftext(base_sizes[[i]],
                  fp_text(font.size = 12, color = "#666666"))
          ),
          location = ph_location(left = 0.5, top = 0.5 + img_height + 0.1,
                                 width = slide_width - 1, height = 0.25)
        )
    }
  }

  print(prs, target = filename)
  unlink(tmp_files)
  message("Saved: ", filename)
}


# plot_stacked() --------------------------------------------------------------
# Aggregates a categorical outcome x predictor using weighted proportions and
# returns a bar_stacked plot.

plot_stacked <- function(outcome, predictor, data) {

  dat <- calc_stats(
    data       = data,
    outcomes   = outcome,
    predictors = list(predictor),
    statistics = "w_perc"
  ) %>%
    filter(cross_break == predictor) %>%
    mutate(estimate = round(estimate * 100, 1))

  title   <- dat$o_lab  %>% unique()
  x_lab   <- dat$p_lab1 %>% unique()

  plt <- bar_stacked(
    dat             = dat,
    x               = p_cat1,
    y               = estimate,
    fill            = o_cat,
    x_lab           = x_lab,
    y_lab           = "%",
    fill_lab        = NULL,
    title           = title,
    base_size       = pptx_base_size,
    legend_position = pptx_legend_position
  )

  list(plot = plt, base_size = get_base_string(dat))

}


# plot_median() ---------------------------------------------------------------
# Aggregates a numeric outcome x predictor using weighted median and returns
# a bar plot.

plot_median <- function(outcome, predictor, data) {

  dat <- calc_stats(
    data       = data,
    outcomes   = outcome,
    predictors = list(predictor),
    statistics = "w_median"
  ) %>%
    filter(cross_break == predictor)

  title <- dat$o_lab  %>% unique()
  x_lab <- dat$p_lab1 %>% unique()

  plt <- bar(
    dat       = dat,
    x         = p_cat1,
    y         = estimate,
    x_lab     = x_lab,
    y_lab     = "Median",
    title     = title,
    base_size = pptx_base_size
  )

  list(plot = plt, base_size = get_base_string(dat))

}
```

<br>

## Plots

### Categorical outcomes

```{r plots-cat, fig.height=4, fig.width=7}

results_cat <- map(outcomes_cat, plot_stacked, predictor = crosstab, data = df_svy)
walk(results_cat, \(x) print(x$plot))
```

<br>

### Numeric outcomes

```{r plots-num, fig.height=4, fig.width=7}

results_num <- map(outcomes_num, plot_median, predictor = crosstab, data = df_svy)
walk(results_num, \(x) print(x$plot))
```

<br>

## Export

```{r export}

all_results <- c(results_cat, results_num)

export_plots_to_pptx(
  plot_list  = map(all_results, "plot"),
  base_sizes = map(all_results, "base_size"),
  filename   = here::here(paste0("7a_Outcomes by ", crosstab, ".pptx"))
)
```
