---
title: "Basic bar charts"
format: html
editor: visual
---

```{r setup}
#| message: false
library(tidyverse)
library(visu)
```

### How bar() works

`bar()` plots statistics already calculated and sitting in a data frame — one row per bar, an x category paired with an already-computed y value (`geom_col()`'s job) — rather than counting raw observations itself (`geom_bar()`'s job). `dat` needs to already be aggregated (e.g. via `count()` or `summarise()`) before it's passed in.

-   `dat` — the data frame, one row per bar
-   `x` — unquoted column name for the x-axis category
-   `y` — unquoted column name for the already-computed value each bar shows
-   `x_lab` — optional x-axis title
-   `y_lab` — optional y-axis title
-   `title` — optional plot title
-   `fill_family` — bar colour: an `nc_alt1` name (`"pink"`, `"blue3"`), `"multi"` (one colour per bar, spread across families), or `"<family>_seq"` (one shade per bar within a single family — ordered x variables only, max 5 bars)
-   `base_size` — base font size in pt, scales all text via `theme_house()`
-   `horiz` — `FALSE` (default) for upright bars, `TRUE` to flip horizontal. Whichever of `x_lab`/`y_lab` ends up on the narrow left margin gets rotated vertical and wrapped once it's long enough (3+ characters) — that's always `y_lab` when upright, but swaps to `x_lab` once flipped (see below)
-   `plot_width` — feeds the label-wrapping maths only, not the plot's actual rendered size (see below)
-   `plot_height` — feeds the wrapping maths for whichever label is on the left margin (see below)

#### plot_width and plot_height, in more detail

They feed `wrap_text()`/`resolve_y_lab()`'s wrapping calculations only — how many characters fit before the x-axis labels, title, or a rotated y-axis title get wrapped — not the plot's actual rendered size. They only matter once something's long enough to need wrapping in the first place; short labels and titles won't react to either. If this were exported via knitting, the chunk's own `fig-width`/`fig-height` options would need to be kept in sync with these by hand — though options exist (e.g. `knitr::opts_current$get("fig.width")`) to link the two automatically, not yet set up here.

### Examples

Same data as `2_geom_gallery.qmd`'s `geom_col` example — one row per class, so `bar()` can be compared directly against the plain `geom_col(fill = "steelblue")` version there.

```{r}
mpg_class <- mpg %>% count(class)
```

### No colour palette

Default `fill_family` — `bar()`'s own default, `"pink"`.

```{r}
bar(mpg_class, class, n)
```

### Bars all the same colour

`fill_family` given the name of a different `nc_alt1` family — still a single colour, applied to every bar.

```{r}
bar(mpg_class, class, n, fill_family = "blue")
```

### A specific shade

`fill_family` also takes a family name with a shade number attached (1-5) — `resolve_pal_colour()` resolves the string, family name alone defaults to shade 1.

```{r}
bar(mpg_class, class, n, fill_family = "mint4")
```

### Bars different colours

`fill_family = "multi"` — `bar()` works out how many bars there are and pulls that many colours from `give_categorical_colours_pal()` internally, one per bar.

```{r}
bar(mpg_class, class, n, fill_family = "multi")
```

### Bars coloured sequentially within one family

`fill_family = "<family>_seq"` — each bar gets a different shade from that one family's tint ramp instead of spreading across families, for when the x variable is ordered (so the colour progression itself carries meaning). Capped at 5 bars, since a family only has 5 shades — `diamonds$cut` is a natural fit, an ordered factor with exactly 5 levels.

```{r}
diamonds_cut <- diamonds %>% count(cut)

bar(diamonds_cut, cut, n, fill_family = "mint_seq", title = "Diamonds by cut")
```

### Custom labels and title

`x_lab`, `y_lab`, and `title` are optional character strings, wrapped to `plot_width`. A short `y_lab` (3 characters or under) stays horizontal.

```{r}
bar(mpg_class, class, n, x_lab = "Vehicle class", y_lab = "Count", title = "Vehicles per class")
```

### Long y-axis label

Past 3 characters, `y_lab` rotates vertical and wraps to `plot_height` instead — `resolve_y_lab()`'s job.

```{r}
bar(mpg_class, class, n, y_lab = "Number of vehicles in the dataset")
```

Narrowing `plot_height` wraps that same label harder.

```{r}
bar(mpg_class, class, n, y_lab = "Number of vehicles in the dataset", plot_height = 2)
```

### base_size

Scales all text through `theme_house()`.

```{r}
bar(mpg_class, class, n, title = "Larger base size", base_size = 16)
```

### More categories

`bar_width_from_n()` widens bars automatically as the category count rises (so they don't thin out to slivers), and past 8 categories the x-axis labels switch from wrapped/horizontal to a single angled line instead — `manufacturer` has 15 levels, several with longer names, a harder test than `class`'s seven short ones.

```{r}
mpg_manufacturer <- mpg %>% count(manufacturer)

bar(mpg_manufacturer, manufacturer, n, fill_family = "multi", title = "Vehicles by manufacturer")
```

### plot_width

`plot_width` only feeds the label-wrapping maths (`wrap_text()`/`resolve_y_lab()`) — it doesn't set how wide the plot actually renders. That's controlled separately, here via the chunk's own `fig-width`, or via `finalise_plot()`'s `ggsave()` width when exporting rather than knitting. Set both together so the wrapping assumption matches the real output.

```{r}
#| fig-width: 3

bar(mpg_manufacturer, manufacturer, n, fill_family = "multi", title = "A much narrower plot", plot_width = 3)
```

### horiz

`horiz = TRUE` flips to horizontal bars. Each category gets its own row rather than competing for horizontal space, so the 45° rotation from the previous section doesn't kick in at all here, even with the same 15-category data.

```{r}
bar(mpg_manufacturer, manufacturer, n, fill_family = "multi", horiz = TRUE, title = "Vehicles by manufacturer")
```

Flipping also swaps which label sits on the narrow left margin — a long `x_lab` now gets the rotate-vertical-and-wrap treatment that `y_lab` got in the "Long y-axis label" section above, rather than `x_lab`'s usual plain wrapping.

```{r}
bar(mpg_manufacturer, manufacturer, n, fill_family = "multi", horiz = TRUE, x_lab = "Manufacturer of vehicle", title = "Vehicles by manufacturer")
```



